---
title: "Working through a real dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working through a real dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
have <- requireNamespace("gapminder", quietly = TRUE) &&
  requireNamespace("palmerpenguins", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", eval = have,
  fig.width = 6.5, fig.height = 4, dpi = 150, out.width = "100%"
)
library(ggplot2)
library(tufter)
```

```{r missing, echo = FALSE, eval = !have, results = "asis"}
cat("This vignette needs the gapminder and palmerpenguins packages.",
    "Install them and rebuild to see the figures.\n")
```

Two datasets, both public domain, both installed from CRAN rather than
downloaded while this page builds. `gapminder` gives life expectancy, income
and population for 142 countries every five years from 1952 to 2007.
`palmerpenguins` gives body measurements for 344 penguins on three islands in
the Palmer Archipelago.

I've used real data here because the awkwardness is the point. Simulated data
is symmetric and well behaved, and most of what these measurements have to say
only shows up when the data isn't.

```{r data}
library(gapminder)
library(palmerpenguins)

gap <- as.data.frame(gapminder)
peng <- as.data.frame(penguins[complete.cases(penguins), ])

dim(gap)
dim(peng)
```

## Start with what ggplot2 gives you

```{r default}
default <- ggplot(peng, aes(flipper_length_mm, body_mass_g)) +
  geom_point(alpha = 0.6, size = 1.2)

default
```

```{r default-ink}
r_default <- data_ink_ratio(default)
r_default
```

The grey panel and the white grid account for most of that, and none of it
changes when a penguin does.

## Erase, then replace the frame

`theme_tufte()` takes off the background, the grid and the border.
`geom_rangeframe()` puts back something the border never gave you: an axis line
that spans only the range the data occupy, so the frame reports the minimum and
maximum for free.

```{r lean}
lean <- default +
  geom_rangeframe() +
  labs(x = "Flipper length (mm)", y = "Body mass (g)") +
  theme_tufte() +
  label_source("Horst, Hill and Gorman, palmerpenguins")

lean
```

```{r lean-ink}
r_lean <- data_ink_ratio(lean)
r_lean
```

`r sprintf("%.1f", r_lean$ratio / r_default$ratio)` times the share of the ink
now varies with the data, on the same 333 penguins.

## A quartile frame, and where it stops working

`geom_quartileframe()` breaks the axis at the five-number summary, so the frame
carries the distribution rather than boxing the panel. `quartile_breaks()` puts
the printed labels in the same places.

```{r quartile}
ggplot(peng, aes(flipper_length_mm, body_mass_g)) +
  geom_point(alpha = 0.6, size = 1.2) +
  geom_quartileframe() +
  scale_x_continuous(breaks = quartile_breaks(peng$flipper_length_mm)) +
  scale_y_continuous(breaks = quartile_breaks(peng$body_mass_g)) +
  labs(x = "Flipper length (mm)", y = "Body mass (g)") +
  theme_tufte()
```

Both variables here are close to symmetric, which is why all five marks sit
comfortably apart. On a skewed variable they crowd at one end. That crowding is
information, since it tells you the distribution is skewed, and it also makes
for an axis nobody can read, so on skewed data I'd use a plain range frame and
let the marks show the skew.

## Distributions: the box plot with the box erased

```{r boxplot, fig.height = 3.2}
ggplot(peng, aes(species, body_mass_g)) +
  geom_tufteboxplot() +
  geom_rangeframe(sides = "l") +
  labs(x = NULL, y = "Body mass (g)") +
  theme_tufte() +
  label_source("palmerpenguins")
```

Gentoo penguins are heavier than the other two by about a kilogram, and the
two lighter species overlap almost completely. The box plot with the box erased
says that in a line and three dots.

## Slopegraphs: a before and after that a bar chart would bury

Life expectancy in South-East Asia, 1952 against 2007. Every number is printed
on the graphic, which makes the y axis redundant, so it goes.

```{r slopegraph, fig.height = 5}
picked <- c("Cambodia", "Indonesia", "Malaysia", "Philippines",
            "Thailand", "Vietnam", "Myanmar")
sea <- subset(gap, country %in% picked & year %in% c(1952, 2007))
sea$year <- factor(sea$year)

slopegraph(sea, year, lifeExp, country, accuracy = 0.1) +
  labs(title = "Life expectancy at birth, 1952 and 2007") +
  label_source("gapminder", note = "Years, South-East Asia.")
```

Every country rose, and the crossings are the interesting part. Thailand began
the highest of the seven and ends fifth. Vietnam began fourth and ends joint
top with Malaysia. Cambodia began third from the bottom and ends last. A pair
of bar charts would have shown exactly the same fourteen numbers and hidden
every one of those reorderings, which is the case for the form.

## Small multiples

`facet_tufte()` fixes the scales, and warns if you try to free them, because
free scales destroy the comparison the design exists to make.

```{r multiples, fig.height = 3.2}
recent <- subset(gap, year >= 1977 & continent != "Oceania")

ggplot(recent, aes(gdpPercap, lifeExp)) +
  geom_point(alpha = 0.25, size = 0.7) +
  geom_rangeframe() +
  facet_tufte(~ continent, ncol = 4) +
  scale_x_log10(labels = scales::label_comma()) +
  labs(x = "GDP per capita (log scale)", y = "Life expectancy") +
  theme_tufte() +
  label_source("gapminder", note = "1977 onward.")
```

## Banking: how tall should the panel be?

One country's life expectancy over 55 years. How steep the rise looks depends
on how tall the panel is, and `bank_to_45()` picks the height that puts the
median slope nearest 45 degrees, where Cleveland found we judge slope most
accurately.

```{r banking}
korea <- subset(gap, country == "Korea, Rep.")

series <- ggplot(korea, aes(year, lifeExp)) +
  geom_line(linewidth = 0.35) +
  geom_rangeframe(sides = "l") +
  labs(x = NULL, y = "Life expectancy") +
  theme_tufte()

bank_to_45(series, width = 6.5)
```

That height is for the panel. A saved figure needs more, for the axis labels
and the y title, and `check_labels_fit()` will tell you when you haven't left
enough.

```{r banked, fig.height = 2.2}
series
```

## Sparklines, when each series has its own scale

```{r sparklines, fig.height = 3}
four <- subset(gap, country %in% c("China", "India", "Japan", "Indonesia"))

sparklines(four, year, gdpPercap, country, accuracy = 1)
```

Four series whose levels differ by an order of magnitude. On shared axes three
of them would flatten against the bottom; each on its own scale, the shapes are
comparable even though the levels aren't. That's the opposite choice from the
small multiples above, and both are honest.

## Dot plots, when zero is a long way away

```{r dotplot, fig.height = 3}
top <- subset(gap, year == 2007 & continent == "Europe")
top <- head(top[order(-top$lifeExp), ], 10)

ggplot(top, aes(lifeExp, stats::reorder(country, lifeExp))) +
  geom_cleveland_dot() +
  labs(x = "Life expectancy in 2007", y = NULL) +
  theme_tufte() +
  label_source("gapminder")
```

These ten countries sit between
`r sprintf("%.1f", min(top$lifeExp))` and `r sprintf("%.1f", max(top$lifeExp))`
years. Bars from zero would put every one at nearly the same length and waste
four fifths of the panel. Bars cropped to the data would make a two-year spread
look like the whole story, and `lie_factor()` puts a number on how much.

```{r lie}
bars <- ggplot(top, aes(stats::reorder(country, lifeExp), lifeExp)) + geom_col()

c(from_zero = lie_factor(bars),
  cropped   = lie_factor(bars + coord_cartesian(ylim = c(79, 82))))
```

A dot encodes its value by position, so it can be read against a scale that
excludes zero without claiming anything about proportions. That's the whole
reason to reach for one.

## Auditing the result

```{r audit}
tufte_audit(lean, width = 6.5, height = 4)
```

## A whole paper at once

```{r batch}
figures <- list(
  `fig 1 penguins` = lean,
  `fig 2 species`  = ggplot(peng, aes(species, body_mass_g)) +
    geom_tufteboxplot() + geom_rangeframe(sides = "l") +
    theme_tufte() + label_source("palmerpenguins"),
  `fig 3 cropped`  = bars + coord_cartesian(ylim = c(79, 81))
)

audit_figures(figures, measure = FALSE)
```

The cropped bar chart comes first because it fails the most stated criteria,
and the count is comparable across the three because every figure is counted
against the same list.
