---
title: "Working through a real dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working through a real dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
have <- requireNamespace("gapminder", quietly = TRUE) &&
  requireNamespace("palmerpenguins", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", eval = have,
  fig.width = 6.5, fig.height = 4, dpi = 300, out.width = "100%"
)
library(ggplot2)
library(tufter)
```

```{r missing, echo = FALSE, eval = !have, results = "asis"}
cat("This vignette needs the gapminder and palmerpenguins packages.",
    "Install them and rebuild to see the figures.\n")
```

Here I use two datasets available in R packages. `gapminder` contains life
expectancy, income and population for 142 countries at five-year intervals
from 1952 to 2007. `palmerpenguins` contains measurements of 344 penguins from
three islands in the Palmer Archipelago.

These examples show where the plotting tools help and where the data call
for a different choice. The page uses the installed datasets and doesn't
download them during the build.

```{r data}
library(gapminder)
library(palmerpenguins)

gap <- as.data.frame(gapminder)
peng <- as.data.frame(penguins[complete.cases(penguins), ])

dim(gap)
dim(peng)
```

## Start with what ggplot2 gives you

```{r default}
default <- ggplot(peng, aes(flipper_length_mm, body_mass_g)) +
  geom_point(alpha = 0.6, size = 1.2)

default
```

```{r default-ink}
r_default <- data_ink_ratio(default)
r_default
```

The panel background and grid account for much of the estimated non-data
ink in this version.

## Erase, then replace the frame

I'll remove those elements with `theme_tufte()` and add a range frame. The
new axis lines show the smallest and largest observed values.

```{r lean}
lean <- default +
  geom_rangeframe() +
  labs(x = "Flipper length (mm)", y = "Body mass (g)") +
  theme_tufte() +
  label_source("Horst, Hill and Gorman, palmerpenguins")

lean
```

```{r lean-ink}
r_lean <- data_ink_ratio(lean)
r_lean
```

The estimated data-ink share is now
`r sprintf("%.1f", r_lean$ratio / r_default$ratio)` times the original share.
Both plots use the same 333 complete observations.

## A quartile frame, and where it stops working

A quartile frame marks the five-number summary. `quartile_breaks()` places
the axis labels at those values.

```{r quartile}
ggplot(peng, aes(flipper_length_mm, body_mass_g)) +
  geom_point(alpha = 0.6, size = 1.2) +
  geom_quartileframe() +
  scale_x_continuous(breaks = quartile_breaks(peng$flipper_length_mm)) +
  scale_y_continuous(breaks = quartile_breaks(peng$body_mass_g)) +
  labs(x = "Flipper length (mm)", y = "Body mass (g)") +
  theme_tufte()
```

The five labels fit here. With a skewed distribution, several can end up
close together. I'd use a plain range frame if the quartile labels became
hard to read.

## Distributions: the box plot with the box erased

```{r boxplot, fig.height = 3.2}
ggplot(peng, aes(species, body_mass_g)) +
  geom_tufteboxplot() +
  geom_rangeframe(sides = "l") +
  labs(x = NULL, y = "Body mass (g)") +
  theme_tufte() +
  label_source("palmerpenguins")
```

Gentoo penguins are heavier on average than the other two species, whose
distributions overlap substantially. The median dots and whiskers make that
comparison visible.

## Slopegraphs: changes in life expectancy

Here's life expectancy in seven South-East Asian countries in 1952 and 2007.
The values are printed at both ends of each line.

```{r slopegraph, fig.height = 5}
picked <- c("Cambodia", "Indonesia", "Malaysia", "Philippines",
            "Thailand", "Vietnam", "Myanmar")
sea <- subset(gap, country %in% picked & year %in% c(1952, 2007))
sea$year <- factor(sea$year)

slopegraph(sea, year, lifeExp, country, accuracy = 0.1) +
  labs(title = "Life expectancy at birth, 1952 and 2007") +
  label_source("gapminder", note = "Years, South-East Asia.")
```

Life expectancy increased in every country shown. The lines also help us
see changes in rank: Thailand starts highest and ends fifth, while Vietnam
moves from fourth to roughly level with Malaysia at the top. Cambodia ends
lowest.

## Small multiples

For the next comparison, I'll keep the same scales across panels.
`facet_tufte()` does that by default and warns if you request free scales.

```{r multiples, fig.height = 3.2}
recent <- subset(gap, year >= 1977 & continent != "Oceania")

ggplot(recent, aes(gdpPercap, lifeExp)) +
  geom_point(alpha = 0.25, size = 0.7) +
  geom_rangeframe() +
  facet_tufte(~ continent, ncol = 4) +
  scale_x_log10(breaks = c(1000, 10000), labels = c("1,000", "10,000")) +
  labs(x = "GDP per capita (log scale)", y = "Life expectancy") +
  theme_tufte() +
  label_source("gapminder", note = "1977 onward.")
```

## Banking: how tall should the panel be?

The apparent steepness of a trend depends on the panel's height. Here I use
`bank_to_45()` to suggest a height for one country's life-expectancy series.
It applies Cleveland's approach to bringing the median absolute slope near
45 degrees.

```{r banking}
korea <- subset(gap, country == "Korea, Rep.")

series <- ggplot(korea, aes(year, lifeExp)) +
  geom_line(linewidth = 0.35) +
  geom_rangeframe(sides = "l") +
  labs(x = NULL, y = "Life expectancy") +
  theme_tufte()

bank_to_45(series, width = 6.5)
```

The suggested height describes the panel. Leave additional space for titles
and axis labels, then run `check_labels_fit()` at the intended export size.

```{r banked, fig.height = 2.2}
series
```

## Sparklines, when each series has its own scale

```{r sparklines, fig.height = 3}
four <- subset(gap, country %in% c("China", "India", "Japan", "Indonesia"))

sparklines(four, year, gdpPercap, country, accuracy = 1)
```

These four series have very different levels. Giving each its own scale
makes the patterns easier to see, but it prevents direct comparisons of
height across series. Use the fixed-scale panels above when levels are the
question.

## Dot plots, when zero is a long way away

```{r dotplot, fig.height = 3}
top <- subset(gap, year == 2007 & continent == "Europe")
top <- head(top[order(-top$lifeExp), ], 10)

ggplot(top, aes(lifeExp, stats::reorder(country, lifeExp))) +
  geom_cleveland_dot() +
  labs(x = "Life expectancy in 2007", y = NULL) +
  theme_tufte() +
  label_source("gapminder")
```

The ten values range from `r sprintf("%.1f", min(top$lifeExp))` to
`r sprintf("%.1f", max(top$lifeExp))` years. Bars starting at zero make the
differences hard to see. Cropping those bars makes their lengths misleading.
Here's what `lie_factor()` reports for the cropped version.

```{r lie}
bars <- ggplot(top, aes(stats::reorder(country, lifeExp), lifeExp)) + geom_col()

c(from_zero = lie_factor(bars),
  cropped   = lie_factor(bars + coord_cartesian(ylim = c(79, 82))))
```

Dots show values by position, so they're useful for this comparison on an
axis that excludes zero.

## Auditing the result

```{r audit}
tufte_audit(lean, width = 6.5, height = 4)
```

## A whole paper at once

```{r batch}
figures <- list(
  `fig 1 penguins` = lean,
  `fig 2 species`  = ggplot(peng, aes(species, body_mass_g)) +
    geom_tufteboxplot() + geom_rangeframe(sides = "l") +
    theme_tufte() + label_source("palmerpenguins"),
  `fig 3 cropped`  = bars + coord_cartesian(ylim = c(79, 81))
)

audit_figures(figures, measure = FALSE)
```

The cropped bar chart appears first because it has the most unmet criteria.
That's a useful place to begin reviewing these figures. Also check whether
any audit steps were skipped.
