---
title: "Designing and measuring a figure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Designing and measuring a figure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6.5, fig.height = 4, dpi = 300, out.width = "100%"
)
library(ggplot2)
library(tufter)
set.seed(1)
```

How can we make a figure easier to read? I built `tufter` to try some of the
ideas in Tufte's *The Visual Display of Quantitative Information* and examine
what changes. It provides plotting tools and estimates of data-ink ratio, lie
factor and data density.

The audit checks explicit criteria and reports the other measurements without
a pass or fail. I've kept those separate because Tufte doesn't give every
principle a numerical target. There's no overall score.

## Start with a default plot

Here's a scatterplot using the defaults in `ggplot2`.

```{r default}
base <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point()

base
```

We can estimate how much of its ink comes from the data layers.

```{r measure-default}
data_ink_ratio(base)
```

The estimate counts the panel background and grid as non-data ink. That
helps explain the low ratio, though it doesn't tell us whether those elements
help a reader.

## Erase, then replace the frame

`theme_tufte()` removes the background, grid and border. Adding
`geom_rangeframe()` draws an axis line between the smallest and largest
observed values.

```{r lean}
lean <- base +
  geom_rangeframe() +
  theme_tufte()

lean
```

```{r measure-lean}
data_ink_ratio(lean)
```

A quartile frame also marks the five-number summary. Pair
`geom_quartileframe()` with `quartile_breaks()` to put labels at those values.

```{r quartile}
base +
  geom_quartileframe() +
  scale_x_continuous(breaks = quartile_breaks(mtcars$wt)) +
  scale_y_continuous(breaks = quartile_breaks(mtcars$mpg)) +
  theme_tufte()
```

## The audit

`tufte_audit()` lists unmet criteria with their sources. It also reports
measurements that you can compare with another version of the figure.

```{r audit}
tufte_audit(lean, width = 6.5, height = 4)
```

This version still needs a data source. Add one with `label_source()`.

```{r audit-documented}
tufte_audit(lean + label_source("Motor Trend, 1974"), width = 6.5, height = 4)
```

The data-ink ratio and data density are reported without grades. I'd compare
them across drafts and ask whether the change makes the figure more useful.
A higher value isn't sufficient reason to prefer a design.

The audit can flag a missing source or a truncated label. It can't assess
whether the figure supports your argument.

## Graphical integrity

Truncating a bar chart's baseline changes the proportions shown by its bars.
`lie_factor()` measures that distortion for supported bar charts.

```{r lie}
d <- data.frame(president = c("Bush", "Obama"), growth = c(100, 110))

honest <- ggplot(d, aes(president, growth)) + geom_col()
truncated <- honest + coord_cartesian(ylim = c(95, 115))

lie_factor(honest)
lie_factor(truncated)
```

Here the displayed proportional change is much larger than the change in
the data. Tufte's reference band is roughly 0.95 to 1.05.

You can also supply numeric vectors, as in the fuel-economy example from
*The Visual Display*. The function compares proportional changes in the
values and in the marks used to draw them.

```{r lie-numeric}
lie_factor(c(18.0, 27.5), c(0.6, 5.3))
```

## Bars with the gridlines erased

Gridlines can help readers estimate bar heights. Tufte's version draws them
as gaps through the bars. Here's how that looks.

```{r bars}
d <- data.frame(
  crop = c("Wheat", "Maize", "Rice", "Barley", "Oats"),
  yield = c(3.5, 5.8, 4.6, 3.1, 2.5)
)

ggplot(d, aes(crop, yield)) +
  geom_col_tufte(fill = "grey72") +
  labs(x = NULL, y = "Tonnes per hectare") +
  theme_tufte()
```

## Box plots without the box

This version uses whisker lines and a median dot, with a gap for the
interquartile range.

```{r boxplot}
ggplot(mtcars, aes(factor(cyl), mpg)) +
  geom_tufteboxplot() +
  geom_rangeframe(sides = "l") +
  labs(x = "Cylinders", y = "Miles per gallon") +
  theme_tufte()
```

Try `type = "line"` or `type = "offset"` if the interquartile range needs a
more visible mark.

## Labels on the data

I often prefer putting a series name beside its line. `geom_text_last()`
labels the final point so readers don't have to look up a legend.

```{r direct}
series <- data.frame(
  year = rep(2000:2015, 3),
  value = c(cumsum(rnorm(16)), cumsum(rnorm(16)) + 4, cumsum(rnorm(16)) - 4),
  crop = rep(c("Wheat", "Maize", "Rice"), each = 16)
)

ggplot(series, aes(year, value, colour = crop)) +
  geom_line(linewidth = 0.4) +
  geom_text_last(aes(label = crop), size = 3) +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.12))) +
  scale_colour_tufte("grey") +
  theme_tufte() +
  theme(legend.position = "none")
```

## Slopegraphs

A slopegraph compares two periods. The lines show changes in values and rank,
and labels give the numbers at each end.

```{r slopegraph, fig.height = 5}
d <- data.frame(
  country = rep(c("Sweden", "Japan", "Chile", "Canada", "Greece"), each = 2),
  year = rep(c("1970", "2020"), 5),
  spending = c(30.1, 41.2, 20.7, 32.9, 22.5, 21.0, 31.0, 38.4, 25.2, 29.7)
)

slopegraph(d, year, spending, country) +
  labs(title = "Public spending as a share of GDP")
```

## Sparklines

Sparklines show a series in a small space. Here the grey band marks the
interquartile range, dots identify the extremes, and a label gives the final
value.

```{r sparklines, fig.height = 2.5}
d <- data.frame(
  month = rep(1:60, 4),
  value = c(cumsum(rnorm(60)), cumsum(rnorm(60)), cumsum(rnorm(60)),
            cumsum(rnorm(60))),
  series = rep(c("Wheat", "Maize", "Rice", "Barley"), each = 60)
)

sparklines(d, month, value, series)
```

## Small multiples

Use one panel per group when you want readers to make the same comparison
several times. `facet_tufte()` keeps the scales fixed so levels can be compared
across panels. It warns if you request free scales.

```{r multiples}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point(size = 1) +
  geom_rangeframe() +
  facet_tufte(~ cyl) +
  theme_tufte()
```

## Before you save

Check the figure at the dimensions you'll use. A long subtitle can fit in a
preview and still be cut off in the saved file. `check_labels_fit()` estimates
the space needed by titles, axes, legends and facet strips. It doesn't check
text placed inside the panel.

```{r fit}
wordy <- lean +
  labs(subtitle = paste(rep("A subtitle that runs on rather too long", 4),
                        collapse = " "))

check_labels_fit(wordy, width = 6.5, height = 4)
```

`save_tufte()` runs that check before saving. It defaults to a width of
6.5 inches and uses `cairo_pdf` for PDF output. Set `strict = TRUE` if a label
that doesn't fit, or a check that can't run, should stop the save.

```{r save, eval = FALSE}
save_tufte("figure-1.pdf", lean, width = 6.5, height = 4)
```

## What the package can't do

`tufte_principles()` lists the principles and their sources. The `audited`
column identifies which ones the package can check.

```{r principles}
p <- tufte_principles()
p[!p$audited, c("principle", "implemented_by")]
```

Whether a figure supports a causal claim or presents a useful comparison
still requires judgement. I use the audit to catch specific problems while
reviewing those larger questions myself.
