---
title: "Designing and measuring a figure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Designing and measuring a figure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6.5, fig.height = 4, dpi = 150, out.width = "100%"
)
library(ggplot2)
library(tufter)
set.seed(1)
```

Tufte's argument in *The Visual Display of Quantitative Information* is that
you can evaluate a statistical graphic. He hands you quantities to do it with:
the data-ink ratio, the lie factor, data density. This package takes him at his
word in both directions. It gives you the forms he designed, and it gives you
the measurements, so a figure you've drawn can be examined as well as
admired.

One distinction runs through all of it. Tufte states a criterion for some
principles, which a graphic either meets or doesn't, and for others he states
only a direction. The audit grades the first kind and measures the second.
There's no score, because collapsing the two would mean making up thresholds
and a weighting that aren't anywhere in his books.

## Start with a default plot

Here is a scatterplot as `ggplot2` draws it out of the box.

```{r default}
base <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point()

base
```

Now measure it.

```{r measure-default}
data_ink_ratio(base)
```

Almost none of that ink is doing any work. The grey panel, the white grid, the
minor gridlines and the axis furniture take up nearly all of it, and none of
them change when the data change.

## Erase, then replace the frame

`theme_tufte()` takes off the background, the grid and the border.
`geom_rangeframe()` gives you something better in place of the border. It draws
the axis line only across the range the data occupy, so the frame reports the
minimum and maximum for free.

```{r lean}
lean <- base +
  geom_rangeframe() +
  theme_tufte()

lean
```

```{r measure-lean}
data_ink_ratio(lean)
```

`geom_quartileframe()` goes further and breaks the axis at the quartiles, so it
carries the whole five-number summary. Pair it with `quartile_breaks()` and the
printed labels will agree with the breaks.

```{r quartile}
base +
  geom_quartileframe() +
  scale_x_continuous(breaks = quartile_breaks(mtcars$wt)) +
  scale_y_continuous(breaks = quartile_breaks(mtcars$mpg)) +
  theme_tufte()
```

## The audit

`tufte_audit()` reports the stated criteria a figure misses, each one named with
the principle and the chapter it comes from. Separately, it measures the
quantities Tufte gives a direction for.

```{r audit}
tufte_audit(lean, width = 6.5, height = 4)
```

The one that's left is the one no theme can fix for you. The figure doesn't say
where its numbers came from. `label_source()` handles that.

```{r audit-documented}
tufte_audit(lean + label_source("Motor Trend, 1974"), width = 6.5, height = 4)
```

Notice what the audit won't do. It never grades the data-ink ratio or the data
density. Tufte asks that both go in a direction and never says how far, so the
audit reports them and leaves the judgement with you. Read them against another
draft of the same figure.

Meeting every stated criterion doesn't make a figure good. Tufte's first
principle is that content counts most of all, and no function evaluates that.
What the audit is good at is the mechanical stuff you stop seeing after the
fifth draft.

## Graphical integrity

The most common way a real figure lies is a bar chart whose baseline isn't zero.
Bar length stops being proportional to the quantity it stands for.
`lie_factor()` measures how far off it gets.

```{r lie}
d <- data.frame(president = c("Bush", "Obama"), growth = c(100, 110))

honest <- ggplot(d, aes(president, growth)) + geom_col()
truncated <- honest + coord_cartesian(ylim = c(95, 115))

lie_factor(honest)
lie_factor(truncated)
```

A ten percent difference drawn as a sixteen-fold one. Tufte treats anything
outside roughly 0.95 to 1.05 as distortion, and I think that's about right.

Given the numbers directly, `lie_factor()` reproduces his own examples. The
fuel-economy graphic in *The Visual Display* showed an eighteen percent change
as a line growing by seven hundred and eighty-three percent.

```{r lie-numeric}
lie_factor(c(18.0, 27.5), c(0.6, 5.3))
```

## Bars with the gridlines erased

Bar charts need gridlines, because readers have to recover values from bar
heights. A gridline crossing a bar, though, is drawn on top of ink that already
carries that value. Tufte's redesign erases it there instead.

```{r bars}
d <- data.frame(
  crop = c("Wheat", "Maize", "Rice", "Barley", "Oats"),
  yield = c(3.5, 5.8, 4.6, 3.1, 2.5)
)

ggplot(d, aes(crop, yield)) +
  geom_col_tufte(fill = "grey72") +
  labs(x = NULL, y = "Tonnes per hectare") +
  theme_tufte()
```

## Box plots without the box

The box in a box plot holds four numbers, and a line and a dot can hold them
just as well.

```{r boxplot}
ggplot(mtcars, aes(factor(cyl), mpg)) +
  geom_tufteboxplot() +
  geom_rangeframe(sides = "l") +
  labs(x = "Cylinders", y = "Miles per gallon") +
  theme_tufte()
```

`type = "line"` and `type = "offset"` keep progressively more, for when the
whiskers are short and the interquartile range needs its own mark.

## Labels on the data, not in a legend

A legend makes the reader look away, hold a colour in memory, look back and
match it up. `geom_text_last()` puts the name where the eye already is.

```{r direct}
series <- data.frame(
  year = rep(2000:2015, 3),
  value = c(cumsum(rnorm(16)), cumsum(rnorm(16)) + 4, cumsum(rnorm(16)) - 4),
  crop = rep(c("Wheat", "Maize", "Rice"), each = 16)
)

ggplot(series, aes(year, value, colour = crop)) +
  geom_line(linewidth = 0.4) +
  geom_text_last(aes(label = crop), size = 3) +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.12))) +
  scale_colour_tufte("grey") +
  theme_tufte() +
  theme(legend.position = "none")
```

## Slopegraphs

A slopegraph shows before-and-after for many units at once. Every number gets
printed on the graphic, which makes the y axis redundant, so it goes. The table
and the figure end up being the same object.

```{r slopegraph, fig.height = 5}
d <- data.frame(
  country = rep(c("Sweden", "Japan", "Chile", "Canada", "Greece"), each = 2),
  year = rep(c("1970", "2020"), 5),
  spending = c(30.1, 41.2, 20.7, 32.9, 22.5, 21.0, 31.0, 38.4, 25.2, 29.7)
)

slopegraph(d, year, spending, country) +
  labs(title = "Public spending as a share of GDP")
```

## Sparklines

A sparkline is word-sized: small enough to sit inside a sentence, with the
normal range as a grey band, dots at the extremes, and the final value printed
at the end.

```{r sparklines, fig.height = 2.5}
d <- data.frame(
  month = rep(1:60, 4),
  value = c(cumsum(rnorm(60)), cumsum(rnorm(60)), cumsum(rnorm(60)),
            cumsum(rnorm(60))),
  series = rep(c("Wheat", "Maize", "Rice", "Barley"), each = 60)
)

sparklines(d, month, value, series)
```

## Small multiples

The answer to multivariate data is repetition rather than complication: the
same graphic, at the same scale, once per condition. `facet_tufte()` fixes the
scales, because free scales destroy the comparison the design exists to make.

```{r multiples}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point(size = 1) +
  geom_rangeframe() +
  facet_tufte(~ cyl) +
  theme_tufte()
```

## Before you save

Design a figure carefully, save it at the wrong size, and you get a truncated
subtitle. `ggplot2` clips long text rather than wrapping it.
`check_labels_fit()` renders at the size you mean to print and measures what
fits.

```{r fit}
wordy <- lean +
  labs(subtitle = paste(rep("A subtitle that runs on rather too long", 4),
                        collapse = " "))

check_labels_fit(wordy, width = 6.5, height = 4)
```

`save_tufte()` runs that check before it writes, so the warning shows up while
you can still do something about it. It defaults to 6.5 inches wide and uses
`cairo_pdf` for PDF output.

```{r save, eval = FALSE}
save_tufte("figure-1.pdf", lean, width = 6.5, height = 4)
```

## What the package can't do

`tufte_principles()` lists every principle, the book it comes from, and the
function that implements it. The `audited` column is the honest part. It marks
which principles a function can check and which ones it can't.

```{r principles}
p <- tufte_principles()
p[!p$audited, c("principle", "implemented_by")]
```

Showing causality, showing comparisons, content counting most of all: a package
can't verify any of that. I think it's what the figure is for.
