---
title: "Designing and measuring a figure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Designing and measuring a figure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6.5, fig.height = 4, dpi = 150, out.width = "100%"
)
library(ggplot2)
library(tufter)
set.seed(1)
```

Tufte's argument in *The Visual Display of Quantitative Information* is that
statistical graphics can be evaluated, not merely preferred. He gives
quantities: the data-ink ratio, the lie factor, data density. This package
takes that seriously in both directions. It supplies the graphical forms he
designed, and it supplies the measurements, so that a figure you have drawn can
be examined rather than admired.

It is careful about one distinction. Tufte states a criterion for some
principles, a graphic either meeting it or not, and only a direction for
others. The audit grades the first kind and measures the second, and there is
no score, because collapsing the two would mean inventing thresholds and a
weighting that are nowhere in his books.

## Start with a default plot

Here is a scatterplot as `ggplot2` draws it out of the box.

```{r default}
base <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point()

base
```

Now measure it.

```{r measure-default}
data_ink_ratio(base)
```

Almost none of that ink is doing any work. The grey panel, the white grid, the
minor gridlines and the axis furniture account for nearly all of it, and none
of them change when the data change.

## Erase, then replace the frame

`theme_tufte()` removes the background, the grid and the border.
`geom_rangeframe()` puts something better in place of the border: an axis line
drawn only across the range the data occupy, so the frame reports the minimum
and maximum for free.

```{r lean}
lean <- base +
  geom_rangeframe() +
  theme_tufte()

lean
```

```{r measure-lean}
data_ink_ratio(lean)
```

`geom_quartileframe()` goes further, breaking the axis at the quartiles so it
carries the whole five-number summary. Pair it with `quartile_breaks()` so the
printed labels agree with the breaks.

```{r quartile}
base +
  geom_quartileframe() +
  scale_x_continuous(breaks = quartile_breaks(mtcars$wt)) +
  scale_y_continuous(breaks = quartile_breaks(mtcars$mpg)) +
  theme_tufte()
```

## The audit

`tufte_audit()` reports the stated criteria a figure does not meet, each named
with the principle and the chapter it comes from, and separately measures the
quantities Tufte gives a direction for.

```{r audit}
tufte_audit(lean, width = 6.5, height = 4)
```

The remaining failure is the one no theme can fix for you: the figure does not
say where its numbers came from. `label_source()` handles that.

```{r audit-documented}
tufte_audit(lean + label_source("Motor Trend, 1974"), width = 6.5, height = 4)
```

Notice what the audit does not do: it never grades the data-ink ratio or the
data density. Tufte asks that both be pushed in a direction and names no
threshold for either, so the audit reports them and leaves the judgement with
you. Read them against another draft of the same figure.

Meeting every stated criterion does not make a figure good. Tufte's first
principle is that content counts most of all, and no function evaluates that.
What the audit is good for is the mechanical failures that authors stop seeing
after the fifth draft.

## Graphical integrity

The most common way a real figure lies is a bar chart whose baseline is not
zero: bar length stops being proportional to the quantity it stands for.
`lie_factor()` measures the distortion.

```{r lie}
d <- data.frame(president = c("Bush", "Obama"), growth = c(100, 110))

honest <- ggplot(d, aes(president, growth)) + geom_col()
truncated <- honest + coord_cartesian(ylim = c(95, 115))

lie_factor(honest)
lie_factor(truncated)
```

A ten percent difference drawn as a sixteen-fold one. Tufte treats anything
outside roughly 0.95 to 1.05 as distortion.

Given the numbers directly, `lie_factor()` reproduces his own examples. The
fuel-economy graphic in *The Visual Display* showed an eighteen percent change
as a line growing by seven hundred and eighty-three percent.

```{r lie-numeric}
lie_factor(c(18.0, 27.5), c(0.6, 5.3))
```

## Bars with the gridlines erased

Bar charts need gridlines, because readers have to recover values from bar
heights. But a gridline crossing a bar is drawn on top of ink that already
encodes the same value. Tufte's redesign erases it there instead.

```{r bars}
d <- data.frame(
  crop = c("Wheat", "Maize", "Rice", "Barley", "Oats"),
  yield = c(3.5, 5.8, 4.6, 3.1, 2.5)
)

ggplot(d, aes(crop, yield)) +
  geom_col_tufte(fill = "grey72") +
  labs(x = NULL, y = "Tonnes per hectare") +
  theme_tufte()
```

## Box plots without the box

The box in a box plot is a container for four numbers that a line and a dot can
hold on their own.

```{r boxplot}
ggplot(mtcars, aes(factor(cyl), mpg)) +
  geom_tufteboxplot() +
  geom_rangeframe(sides = "l") +
  labs(x = "Cylinders", y = "Miles per gallon") +
  theme_tufte()
```

`type = "line"` and `type = "offset"` keep progressively more, for when the
whiskers are short and the interquartile range needs its own mark.

## Labels on the data, not in a legend

A legend makes the reader look away, hold a colour in memory, look back, and
match. `geom_text_last()` puts the name where the eye already is.

```{r direct}
series <- data.frame(
  year = rep(2000:2015, 3),
  value = c(cumsum(rnorm(16)), cumsum(rnorm(16)) + 4, cumsum(rnorm(16)) - 4),
  crop = rep(c("Wheat", "Maize", "Rice"), each = 16)
)

ggplot(series, aes(year, value, colour = crop)) +
  geom_line(linewidth = 0.4) +
  geom_text_last(aes(label = crop), size = 3) +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.12))) +
  scale_colour_tufte("grey") +
  theme_tufte() +
  theme(legend.position = "none")
```

## Slopegraphs

A slopegraph shows before-and-after for many units at once. Every number is
printed on the graphic, which makes the y axis redundant, so it goes. The table
and the figure become the same object.

```{r slopegraph, fig.height = 5}
d <- data.frame(
  country = rep(c("Sweden", "Japan", "Chile", "Canada", "Greece"), each = 2),
  year = rep(c("1970", "2020"), 5),
  spending = c(30.1, 41.2, 20.7, 32.9, 22.5, 21.0, 31.0, 38.4, 25.2, 29.7)
)

slopegraph(d, year, spending, country) +
  labs(title = "Public spending as a share of GDP")
```

## Sparklines

A sparkline is word-sized: small enough to sit inside a sentence, with the
normal range as a grey band, dots at the extremes, and the final value printed
at the end.

```{r sparklines, fig.height = 2.5}
d <- data.frame(
  month = rep(1:60, 4),
  value = c(cumsum(rnorm(60)), cumsum(rnorm(60)), cumsum(rnorm(60)),
            cumsum(rnorm(60))),
  series = rep(c("Wheat", "Maize", "Rice", "Barley"), each = 60)
)

sparklines(d, month, value, series)
```

## Small multiples

The answer to multivariate data is repetition rather than complication: the
same graphic, at the same scale, once per condition. `facet_tufte()` fixes the
scales, because free scales destroy the comparison the design exists to make.

```{r multiples}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point(size = 1) +
  geom_rangeframe() +
  facet_tufte(~ cyl) +
  theme_tufte()
```

## Before you save

A figure designed carefully and then saved at the wrong size is a figure with a
truncated subtitle. `ggplot2` does not wrap long text, it clips it.
`check_labels_fit()` renders at the size you intend to print and measures.

```{r fit}
wordy <- lean +
  labs(subtitle = paste(rep("A subtitle that runs on rather too long", 4),
                        collapse = " "))

check_labels_fit(wordy, width = 6.5, height = 4)
```

`save_tufte()` runs that check before writing, so the warning arrives while you
can still act on it, and defaults to 6.5 inches wide with `cairo_pdf` for PDF
output.

```{r save, eval = FALSE}
save_tufte("figure-1.pdf", lean, width = 6.5, height = 4)
```

## What the package cannot do

`tufte_principles()` lists every principle, the book it comes from, and the
function that implements it. The `audited` column is the honest part: it marks
which principles a function can check and which it cannot.

```{r principles}
p <- tufte_principles()
p[!p$audited, c("principle", "implemented_by")]
```

Showing causality, showing comparisons, and content counting most of all are
not things a package can verify. They are what the figure is for.
