---
title: "Examples with live API data"
description: >
  Every graphical form in the package, drawn from four public APIs that need no
  key: CRAN download logs, USGS earthquakes, Open-Meteo temperatures and
  Wikipedia pageviews.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6.5, fig.height = 4, dpi = 150, out.width = "100%",
  fig.align = "center", dev = "png"
)
library(ggplot2)
library(tufter)
```

The [other gallery](simulated-examples.html) uses simulated data, which is handy
because you can regenerate it anywhere. This one uses real numbers from four
public APIs, none of which need a key or a login:

| Source | What it gives | Endpoint |
|---|---|---|
| CRAN download logs | daily downloads per package | `cranlogs.r-pkg.org` |
| USGS FDSN event service | every earthquake above a magnitude | `earthquake.usgs.gov` |
| Open-Meteo ERA5 archive | daily temperature anywhere, back to 1940 | `archive-api.open-meteo.com` |
| Wikimedia REST API | daily pageviews per article | `wikimedia.org` |

I picked these because real data misbehaves in ways simulated data doesn't, and
the measurements in this package are more interesting when they're pointed at
something messy.

## How the fetching works

The article reads a cached snapshot rather than calling the APIs while the page
builds. Four API calls on every documentation build would be slow, impolite to
whoever runs them, and would break the build on the day one of them is down.
The fetching lives in
[`data-raw/fetch_live_examples.R`](https://github.com/lobsterbush/tufter/blob/main/data-raw/fetch_live_examples.R),
which you can run yourself to refresh everything.

Here's the shape of it. Each call is one URL and one `fromJSON()`, with no
authentication anywhere.

```{r fetch-code, eval = FALSE}
library(jsonlite)

# Daily downloads for one package, last 180 days.
fromJSON(sprintf(
  "https://cranlogs.r-pkg.org/downloads/daily/%s:%s/ggplot2",
  Sys.Date() - 182, Sys.Date() - 2
))$downloads[[1]]

# Every earthquake at magnitude 4.5 or above in the last year.
fromJSON(paste0(
  "https://earthquake.usgs.gov/fdsnws/event/1/query?format=geojson",
  "&starttime=", Sys.Date() - 365, "&endtime=", Sys.Date(),
  "&minmagnitude=4.5&orderby=time"
))

# Daily mean temperature at a point, for a decade.
fromJSON(paste0(
  "https://archive-api.open-meteo.com/v1/archive?latitude=35.68&longitude=139.69",
  "&start_date=2015-01-01&end_date=2024-12-31",
  "&daily=temperature_2m_mean&timezone=UTC"
))$daily

# Daily pageviews for one Wikipedia article.
fromJSON(paste0(
  "https://wikimedia.org/api/rest_v1/metrics/pageviews/per-article",
  "/en.wikipedia/all-access/all-agents/Edward_Tufte/daily/20260401/20260731"
))$items
```

The snapshot ships with the package, so everything below runs offline.

```{r load}
live <- readRDS(system.file("extdata", "live-examples.rds", package = "tufter"))

format(live$fetched_at, "%d %B %Y")
vapply(live[c("cran", "quakes", "climate", "pageviews")], nrow, integer(1))
```

## Sparklines: CRAN downloads

Five packages, daily downloads, six months each. Sparklines are the right form
here because each series has its own scale and I only want the shape. `targets`
gets a few hundred downloads a day and `ggplot2` gets tens of thousands, so
plotting them on shared axes would flatten four of the five into a line along
the bottom.

The grey band is the interquartile range, the blue and red dots are the minimum
and maximum, and the number at the right is the last value. `accuracy = 1`
rounds those to whole downloads, since a tenth of a download would be an odd
thing to print.

```{r cran-sparklines, fig.height = 3}
sparklines(live$cran, day, downloads, package, accuracy = 1)
```

The spike in `ggplot2` is real. It hits `r format(max(live$cran$downloads[live$cran$package == "ggplot2"]), big.mark = ",")`
downloads in a day against a median of
`r format(median(live$cran$downloads[live$cran$package == "ggplot2"]), big.mark = ",")`,
which is the kind of thing you notice in a sparkline and would miss in a table.
I don't know what caused it. A release, a mirror re-syncing, or a large CI fleet
would all look like this from the outside.

## Banking: how tall should that panel be?

Take one of those series on its own. The shape you see depends on how tall you
draw it, and `bank_to_45()` picks the height that puts the median slope nearest
45 degrees, where Cleveland found we judge slope most accurately.

```{r banking}
gg <- subset(live$cran, package == "ggplot2")

series <- ggplot(gg, aes(day, downloads)) +
  geom_line(linewidth = 0.3) +
  geom_rangeframe(sides = "l") +
  labs(x = NULL, y = "Daily downloads") +
  theme_tufte()

bank_to_45(series, width = 6.5)
```

That height is for the panel, and a saved figure needs room on top of it for
the axis labels and the y title. I found this out the hard way while writing
this page: at the banked height exactly, "Daily downloads" runs off the top of
the figure. `check_labels_fit()` now catches it, and it didn't before, which is
the second bug this article turned up.

```{r banked-check}
check_labels_fit(series, width = 6.5, height = 1.3)
```

So give it a little more room. The panel is still close to banked and the title
now fits.

```{r banked, fig.height = 2.1}
series
```

The weekly cycle is the obvious feature once the panel is short. Downloads fall
at weekends, which you can see as a regular beat rather than having to be told.
The three tall spikes and the early plateau stay visible, which is the point of
banking: the routine variation reads as slope instead of as a solid band, so the
unusual days stand out against it.

## Slopegraphs: two decades of temperature

Six cities around the Pacific, mean daily temperature averaged over 1975 to 1984
and again over 2015 to 2024. Every number gets printed on the graphic, so the y
axis is redundant and goes.

```{r slopegraph, fig.height = 5}
slopegraph(live$climate, period, mean_temp, city, accuracy = 0.1) +
  labs(
    title = "Mean daily temperature, two ten-year windows",
    subtitle = "Degrees Celsius, ERA5 reanalysis at a single grid point per city"
  ) +
  label_source("Open-Meteo ERA5 archive", note = "Fetched 2 August 2026.")
```

All six rise, by between
`r sprintf("%.1f", min(with(live$climate, tapply(mean_temp, city, diff))))` and
`r sprintf("%.1f", max(with(live$climate, tapply(mean_temp, city, diff))))`
degrees. I want to be careful about what that does and doesn't show. These are
single grid points from a reanalysis product, so a city with a lot of building
between the two windows picks up its own heat island along with everything else,
and two decades isn't long enough to separate the two. It's a real comparison of
two real averages, and it isn't a climate attribution.

What the slopegraph does well here is the ordering and the one place it changes.
Singapore is the warmest and moved least, at 0.4 degrees. Tokyo started
second-coolest of the six and moved most, and it's the only line that crosses
another: it begins below Melbourne and ends above it. That crossing is the sort
of thing a slopegraph is for, and the sort of thing a pair of bar charts would
bury.

## Distributions: earthquakes

`r format(nrow(live$quakes), big.mark = ",")` earthquakes at magnitude 4.5 or
above in the last year, with magnitude against depth. A quartile frame breaks
the axis at the five-number summary, so the frame reports the distribution
instead of boxing the panel.

```{r quakes-setup}
q <- live$quakes
```

I've used one on magnitude and a plain range frame on depth, and the reason is
worth spelling out. Depth is badly skewed: the median is
`r round(median(q$depth_km))` km against a maximum of `r round(max(q$depth_km))`,
so four of its five quartile marks pile up in the first tenth of the axis and
the labels sit on top of each other. The crowding is real information, since
it's telling you the distribution is skewed, and it also makes for an axis
nobody can read. On a variable like that I'd rather use a range frame and let
the points show the skew.

```{r quakes-frame}
ggplot(q, aes(depth_km, magnitude)) +
  geom_point(alpha = 0.15, size = 0.7) +
  geom_quartileframe(sides = "l") +
  geom_rangeframe(sides = "b") +
  scale_y_continuous(breaks = quartile_breaks(q$magnitude)) +
  labs(x = "Depth (km)", y = "Magnitude") +
  theme_tufte() +
  label_source("USGS FDSN event service", note = "Magnitude 4.5 and above.")
```

Magnitude is skewed too, and there the frame handles it well. The breaks sit at
`r paste(quartile_breaks(q$magnitude)(range(q$magnitude)), collapse = ", ")`,
so three quarters of the events fall in the bottom sixth of the axis and the
long empty stretch above 5 is the whole story about how rare big earthquakes
are.

Split by the standard depth classes and the minimal box plot does the
comparison.

```{r quakes-box, fig.height = 3.2}
q$zone <- cut(
  q$depth_km, c(-Inf, 70, 300, Inf),
  labels = c("Shallow\n(<70 km)", "Intermediate\n(70-300 km)", "Deep\n(>300 km)")
)

ggplot(q, aes(zone, magnitude)) +
  geom_tufteboxplot(outliers = FALSE) +
  geom_rangeframe(sides = "l") +
  labs(x = NULL, y = "Magnitude") +
  theme_tufte() +
  label_source("USGS FDSN event service", note = "Magnitude 4.5 and above.")
```

Two things to say about that figure, both of which took me a moment.

The first is that the three distributions are nearly identical. Deep events are
much rarer, `r sum(q$depth_km > 300)` against
`r format(sum(q$depth_km <= 70), big.mark = ",")` shallow ones, and their median
magnitude is `r median(q$magnitude[q$depth_km > 300])` against
`r median(q$magnitude[q$depth_km <= 70])` for shallow. I'd not make anything of
a tenth of a magnitude unit in a catalogue that's been censored at 4.5, and I
mention it mostly because I'd assumed going in that deep earthquakes would be
larger. They aren't, at least not here.

The second is that I've turned the outlying points off, which needs defending in
a package that keeps telling you outliers are data. Earthquake magnitudes follow
roughly an exponential distribution, so the usual rule of 1.5 times the
interquartile range flags between four and eleven percent of events depending on
the class. Those aren't anomalies, they're the ordinary tail, and drawing three
hundred of them buries the summary the box plot exists to show. The scatterplot
above already shows every one of them.

The lower whiskers are worth reading carefully too. They stop at 4.5 because
that's what I asked the API for, so they're an artefact of the query rather than
of the earth. That belongs on the figure rather than in a reader's head, which
is what the note in `label_source()` is doing.

## Dot plots: Wikipedia pageviews

Five articles, total views over four months. A dot plot suits this better than
bars, because zero is a long way below the smallest value and starting there
would waste most of the panel. A dot encodes its value by position, so it reads
honestly on a scale that excludes zero.

```{r pageviews-dot, fig.height = 2.6}
totals <- aggregate(views ~ article, live$pageviews, sum)

ggplot(totals, aes(views, stats::reorder(article, views))) +
  geom_cleveland_dot() +
  scale_x_continuous(labels = scales::label_comma()) +
  labs(x = "Total pageviews", y = NULL) +
  theme_tufte() +
  label_source("Wikimedia REST API", note = "English Wikipedia, all agents.")
```

Histogram beats the other four put together, which I didn't expect and which I
think says more about homework than about interest in graphical design.

Sort before you plot. An alphabetical dot plot throws away the main advantage of
the form, which is that you can read rank straight off it.

## Small multiples

The same five series again, this time at a fixed scale so the levels are
comparable. That's what small multiples are for, and it's the opposite choice
from the sparklines at the top of this page.

```{r pageviews-multiples, fig.height = 3.5}
ggplot(live$pageviews, aes(date, views)) +
  geom_line(linewidth = 0.25, colour = "grey30") +
  geom_rangeframe(sides = "l") +
  facet_tufte(~ article, ncol = 5) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(x = NULL, y = "Daily views") +
  theme_tufte() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  label_source("Wikimedia REST API")
```

At a fixed scale you can see that four of the five are quiet and one isn't. The
sparklines told you the shape of each series and hid that comparison; this tells
you the comparison and hides most of the shape. Both are honest and they answer
different questions.

## Auditing a real figure

Everything above can be measured. Here's the earthquake figure, audited at the
size I'd print it.

```{r audit}
quake_figure <- ggplot(q, aes(depth_km, magnitude)) +
  geom_point(alpha = 0.15, size = 0.7) +
  geom_quartileframe() +
  labs(x = "Depth (km)", y = "Magnitude") +
  theme_tufte() +
  label_source("USGS FDSN event service")

tufte_audit(quake_figure, width = 6.5, height = 4)
```

```{r audit-measures, include = FALSE}
qi <- data_ink_ratio(quake_figure, width = 6.5, height = 4)
qd <- data_density(quake_figure, width = 6.5, height = 4)
```

Both measurements are worth a look. The data density comes out at
`r round(qd$density)` numbers per square inch, against about three for the
mtcars scatterplot in the [measuring article](measuring.html).
`r format(qd$rows, big.mark = ",")` events over `r qd$variables` mapped
variables will do that, and comparing figures against each other is exactly
what Tufte uses this quantity for. He never sets a minimum.

The data-ink ratio comes out at `r sprintf("%.2f", qi$ratio)`, far above
anything else in these docs, which makes sense once you look at what's on the
page. Eight thousand marks is a lot of ink, and the frame it sits inside is
three hairlines and some text. If anything the true figure is higher still,
because overlapping points get counted once and this plot has a great many
piled up near the origin. The measuring article works through that bias.

Those two numbers are written into the sentence by inline R code rather than
typed, which is the habit I'd recommend for any figure whose values appear in
the surrounding prose. Refresh the data and the sentence refreshes with it.

## Contrast, on a figure drawn faintly on purpose

Those points are drawn at 15 percent opacity, which is the right call for
overplotting and a bad one for anyone reading the figure on a projector.
`check_contrast()` composites the transparency first, so it measures what the
reader actually sees.

```{r contrast}
check_contrast(quake_figure)
```

Black at 15 percent opacity composites to `#D8D8D8`, which comes out at 1.4
against white and sits well below the minimum of 3. In a scatterplot I think
that's a defensible trade, since the shape of the cloud carries the message and
no individual point matters. I'd still rather know than not. If this were going
on a slide I'd raise the alpha and shrink the points instead.

## Refreshing the data

```r
Rscript data-raw/fetch_live_examples.R
```

That rewrites `inst/extdata/live-examples.rds` and this page picks up the new
numbers on the next build. The prose above quotes specific values, so if you
refresh and the numbers move, the sentences around them need a read as well.
That's the same problem as any paper with numbers in the text, and the same fix:
pipe the values in from the data rather than typing them, which is what the
inline code in this article does.
