---
title: "Examples with live API data"
description: >
  Examples using saved API data on package downloads, earthquakes,
  temperatures and pageviews.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6.5, fig.height = 4, dpi = 300, out.width = "100%",
  fig.align = "center", dev = "png"
)
library(ggplot2)
library(tufter)
```

Here I use a saved snapshot from four public APIs. The
[simulated examples](simulated-examples.html) are useful for controlled
comparisons; these data show how the tools handle less tidy distributions.

| Source | What it gives | Endpoint |
|---|---|---|
| CRAN download logs | daily downloads per package | `cranlogs.r-pkg.org` |
| USGS FDSN event service | every earthquake above a magnitude | `earthquake.usgs.gov` |
| Open-Meteo ERA5 archive | daily temperature anywhere, back to 1940 | `archive-api.open-meteo.com` |
| Wikimedia REST API | daily pageviews per article | `wikimedia.org` |

## How the fetching works

The page reads the saved snapshot, so a documentation build doesn't depend
on the APIs being available. To refresh it, run
[`data-raw/fetch_live_examples.R`](https://github.com/lobsterbush/tufter/blob/main/data-raw/fetch_live_examples.R).
The APIs used here don't require authentication.

Here's an example of the requests.

```{r fetch-code, eval = FALSE}
library(jsonlite)

# Daily downloads for one package, last 180 days.
fromJSON(sprintf(
  "https://cranlogs.r-pkg.org/downloads/daily/%s:%s/ggplot2",
  Sys.Date() - 182, Sys.Date() - 2
))$downloads[[1]]

# Every earthquake at magnitude 4.5 or above in the last year.
fromJSON(paste0(
  "https://earthquake.usgs.gov/fdsnws/event/1/query?format=geojson",
  "&starttime=", Sys.Date() - 365, "&endtime=", Sys.Date(),
  "&minmagnitude=4.5&orderby=time"
))

# Daily mean temperature at a point, for a decade.
fromJSON(paste0(
  "https://archive-api.open-meteo.com/v1/archive?latitude=35.68&longitude=139.69",
  "&start_date=2015-01-01&end_date=2024-12-31",
  "&daily=temperature_2m_mean&timezone=UTC"
))$daily

# Daily pageviews for one Wikipedia article.
fromJSON(paste0(
  "https://wikimedia.org/api/rest_v1/metrics/pageviews/per-article",
  "/en.wikipedia/all-access/all-agents/Edward_Tufte/daily/20260401/20260731"
))$items
```

The cached snapshot is included in the repository, so these examples can
be built offline once the required packages are installed.

```{r load}
live <- readRDS("../../data-raw/live-examples.rds")

format(live$fetched_at, "%d %B %Y")
vapply(live[c("cran", "quakes", "climate", "pageviews")], nrow, integer(1))
```

## Sparklines: CRAN downloads

These are six months of daily downloads for five packages. I've given each
series its own scale so its pattern is visible. That means heights can't be
compared across rows.

The grey band marks the interquartile range. Dots identify the extremes and
the label gives the final value. `accuracy = 1` rounds downloads to whole
numbers.

```{r cran-sparklines, fig.height = 3}
sparklines(live$cran, day, downloads, package, accuracy = 1)
```

The largest `ggplot2` spike reaches
`r format(max(live$cran$downloads[live$cran$package == "ggplot2"]), big.mark = ",")`
downloads in a day, compared with a median of
`r format(median(live$cran$downloads[live$cran$package == "ggplot2"]), big.mark = ",")`.
I don't know what caused it. These counts alone can't distinguish a change
in demand from automated downloading.

## Banking: how tall should that panel be?

For a single series, `bank_to_45()` suggests a panel height using Cleveland's
approach to bringing the median absolute slope near 45 degrees.

```{r banking}
gg <- subset(live$cran, package == "ggplot2")

series <- ggplot(gg, aes(day, downloads)) +
  geom_line(linewidth = 0.3) +
  geom_rangeframe(sides = "l") +
  labs(x = NULL, y = "Daily downloads") +
  theme_tufte()

bank_to_45(series, width = 6.5)
```

The suggested height is for the panel. Titles and axis labels need extra
space, which the following label check helps estimate.

```{r banked-check}
check_labels_fit(series, width = 6.5, height = 1.3)
```

I'll add some height to give the labels room, then inspect the figure.

```{r banked, fig.height = 2.1}
series
```

The shorter panel makes the recurring variation easy to see while keeping
the large spikes visible. I'd compare this with a taller version if the
individual daily changes were the focus.

## Slopegraphs: two decades of temperature

This compares mean daily temperature in six Pacific-region cities for
1975 to 1984 and 2015 to 2024. Each endpoint is a ten-year average.

```{r slopegraph, fig.height = 5}
slopegraph(live$climate, period, mean_temp, city, accuracy = 0.1) +
  labs(
    title = "Mean daily temperature, two ten-year windows",
    subtitle = "Degrees Celsius, ERA5 reanalysis at a single grid point per city"
  ) +
  label_source("Open-Meteo ERA5 archive", note = "Fetched 2 August 2026.")
```

All six averages increase, by between
`r sprintf("%.1f", min(with(live$climate, tapply(mean_temp, city, diff))))` and
`r sprintf("%.1f", max(with(live$climate, tapply(mean_temp, city, diff))))`
degrees. These are selected grid points from a reanalysis product. The
comparison doesn't identify the causes of the changes.

The crossing between Tokyo and Melbourne is easy to see: Tokyo starts below
Melbourne and ends above it. Singapore has the highest average in both periods.

## Distributions: earthquakes

The snapshot contains `r format(nrow(live$quakes), big.mark = ",")` earthquakes
at magnitude 4.5 or above during its one-year collection window. Here's
magnitude against depth, with a quartile frame on magnitude.

```{r quakes-setup}
q <- live$quakes
```

I've used a plain range frame on depth because the distribution is strongly
skewed. The median is `r round(median(q$depth_km))` km and the maximum is
`r round(max(q$depth_km))` km. Several quartile labels would crowd together.
The points still show the concentration of shallow events.

```{r quakes-frame}
ggplot(q, aes(depth_km, magnitude)) +
  geom_point(alpha = 0.15, size = 0.7) +
  geom_quartileframe(sides = "l") +
  geom_rangeframe(sides = "b") +
  scale_y_continuous(breaks = quartile_breaks(q$magnitude)) +
  labs(x = "Depth (km)", y = "Magnitude") +
  theme_tufte() +
  label_source("USGS FDSN event service", note = "Magnitude 4.5 and above.")
```

The magnitude breaks are
`r paste(quartile_breaks(q$magnitude)(range(q$magnitude)), collapse = ", ")`.
Here the labels have enough room. Most events are near the lower end of the
recorded range.

Next I'll compare magnitude across depth classes using minimal box plots.

```{r quakes-box, fig.height = 3.2}
q$zone <- cut(
  q$depth_km, c(-Inf, 70, 300, Inf),
  labels = c("Shallow\n(<70 km)", "Intermediate\n(70-300 km)", "Deep\n(>300 km)")
)

ggplot(q, aes(zone, magnitude)) +
  geom_tufteboxplot(outliers = FALSE) +
  geom_rangeframe(sides = "l") +
  labs(x = NULL, y = "Magnitude") +
  theme_tufte() +
  label_source("USGS FDSN event service", note = "Magnitude 4.5 and above.")
```

There are `r sum(q$depth_km > 300)` deep events and
`r format(sum(q$depth_km <= 70), big.mark = ",")` shallow events. Their median
magnitudes are `r median(q$magnitude[q$depth_km > 300])` and
`r median(q$magnitude[q$depth_km <= 70])`, respectively. Those differences
should be read in the context of the catalogue's 4.5-magnitude cutoff.

I've hidden the outlying points in this summary because the scatterplot
above already shows them. The lower whiskers stop at the query cutoff;
that limit is stated in the figure's source note.

## Dot plots: Wikipedia pageviews

These are total views of five articles over the snapshot's four-month
window. A dot plot lets us compare their positions without using bars that
start far below the smallest value.

```{r pageviews-dot, fig.height = 2.6}
totals <- aggregate(views ~ article, live$pageviews, sum)

ggplot(totals, aes(views, stats::reorder(article, views))) +
  geom_cleveland_dot() +
  scale_x_continuous(labels = scales::label_comma()) +
  labs(x = "Total pageviews", y = NULL) +
  theme_tufte() +
  label_source("Wikimedia REST API", note = "English Wikipedia, all agents.")
```

The Histogram article has the most views. I don't know what explains the
difference. Sorting by the totals makes the ordering easy to read.

## Small multiples

Here are the same series over time, with a fixed scale across panels so
readers can compare their levels.

```{r pageviews-multiples, fig.height = 3.5}
ggplot(live$pageviews, aes(date, views)) +
  geom_line(linewidth = 0.25, colour = "grey30") +
  geom_rangeframe(sides = "l") +
  facet_tufte(~ article, ncol = 5, labeller = label_wrap_gen(width = 12)) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(x = NULL, y = "Daily views") +
  theme_tufte() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  label_source("Wikimedia REST API")
```

The shared scale makes differences in levels visible, though smaller
fluctuations are harder to see. The sparklines above are more useful for
examining each series' pattern.

## Auditing a real figure

Here's the earthquake figure checked at the size I'd print it.

```{r audit}
quake_figure <- ggplot(q, aes(depth_km, magnitude)) +
  geom_point(alpha = 0.15, size = 0.7) +
  geom_quartileframe() +
  labs(x = "Depth (km)", y = "Magnitude") +
  theme_tufte() +
  label_source("USGS FDSN event service")

tufte_audit(quake_figure, width = 6.5, height = 4)
```

```{r audit-measures, include = FALSE}
qi <- data_ink_ratio(quake_figure, width = 6.5, height = 4)
qd <- data_density(quake_figure, width = 6.5, height = 4)
```

The estimated data density is `r round(qd$density)` entries per square inch,
using `r format(qd$rows, big.mark = ",")` rows and `r qd$variables` mapped
variables. The [measuring article](measuring.html) explains how those counts
are constructed.

The estimated data-ink ratio is `r sprintf("%.2f", qi$ratio)`. Many points
overlap here, so their combined ink is counted once. I'd interpret the ratio
with that limit in mind.

## Contrast, on a figure drawn faintly on purpose

The points use 15 percent opacity to show where observations overlap.
`check_contrast()` checks the composited colour of an individual mark
against the background. It doesn't model the darker areas created by
overlapping marks.

```{r contrast}
check_contrast(quake_figure)
```

An isolated black point at this opacity has a contrast ratio of about 1.4
against white, below the default minimum of 3. That makes individual points
hard to distinguish. I'd increase opacity if identifying those points were
important, and check a projected version separately.

## Refreshing the data

```r
Rscript data-raw/fetch_live_examples.R
```

The script refreshes `data-raw/live-examples.rds`. Rebuild this page to use
the new snapshot, then inspect the figures and their interpretation again.
Values inserted with inline R update automatically.
