---
title: "Measuring a figure"
description: >
  What the data-ink ratio, the lie factor and data density actually compute,
  and where each of them will mislead you.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6.5, fig.height = 4, dpi = 150, out.width = "100%",
  fig.align = "center", dev = "png"
)
library(ggplot2)
library(tufter)
```

Most of what people take from Tufte is a look. I think the more interesting
claim in *The Visual Display of Quantitative Information* is methodological.
He says you can evaluate a statistical graphic, and he gives you the quantities
to do it with. This article works through what each of those means here, how
it's computed, and where it'll lead you astray if you quote it without
thinking.

All data below are simulated.

```{r data}
set.seed(20260729)
n <- 400
d <- data.frame(
  x = rnorm(n, 50, 12),
  g = sample(c("Treated", "Control"), n, replace = TRUE)
)
d$y <- 12 + 0.42 * d$x + ifelse(d$g == "Treated", 3.1, 0) + rnorm(n, 0, 5)
```

## The data-ink ratio

Tufte defines the data-ink ratio as the share of a graphic's ink given over to
the non-redundant display of data, and asks that it be pushed towards one.
That's a definition without a procedure attached. No book tells you how to
count ink on a screen.

`data_ink_ratio()` estimates it empirically. It renders the plot to a bitmap
twice, once whole and once with every data layer stripped out of the panel
grobs. Each pixel counts in proportion to how far it sits from the background
colour, so an anti-aliased edge counts as the fraction of a pixel it really is.
Subtract, and what's left is data ink.

```{r ratio-default}
base <- ggplot(d, aes(x, y)) +
  geom_point(size = 1, alpha = 0.5) +
  labs(x = "Pre-treatment score", y = "Outcome")

r_base <- data_ink_ratio(base)
r_base
```

Four fifths of that ink is furniture. The grey panel alone is most of it, and
it doesn't change when the data change. Four hundred semi-transparent points
still put a fair amount of ink on the page, which is why the figure isn't as
lopsided as the forty-point scatterplot in the
[getting started vignette](tufter.html).

```{r ratio-lean}
lean <- base + geom_rangeframe() + theme_tufte()
r_lean <- data_ink_ratio(lean)
r_lean
```

The same numbers, with `r sprintf("%.1f", r_lean$ratio / r_base$ratio)` times
the share of the ink doing work. That multiple is written in by inline code
rather than typed, because it moves whenever the measurement does, and it has
already moved once while I was writing this page.

### Where it misleads

Three things are worth knowing before you quote the number.

Overlap gets counted once. A dense scatterplot draws many points on top of each
other and the measurement sees one blob, so dense plots understate their own
data-ink. That's the opposite of the direction you'd want the bias to run.

```{r overlap, fig.height = 3}
dense <- data.frame(x = rnorm(20000), y = rnorm(20000))
sparse <- dense[1:60, ]

lean_theme <- list(geom_rangeframe(), theme_tufte())

c(
  dense = data_ink_ratio(
    ggplot(dense, aes(x, y)) + geom_point(size = 0.4, alpha = 0.2) + lean_theme
  )$ratio,
  sparse = data_ink_ratio(
    ggplot(sparse, aes(x, y)) + geom_point(size = 0.4) + lean_theme
  )$ratio
)
```

Redundant data-ink still counts as data-ink. Tufte would subtract ink that
repeats what the reader already has. No measurement can tell whether a mark is
repeating something, so this one doesn't try.

The number depends on the size you render at. Non-data ink is mostly fixed
furniture, so shrinking the canvas raises the ratio. Compare figures at the size
you mean to print them, and compare like with like.

```{r size-dependence}
vapply(
  c(3, 6.5, 12),
  function(w) data_ink_ratio(lean, width = w, height = w * 0.6)$ratio,
  numeric(1)
)
```

Treat it as a comparative instrument. I think it's reliable for judging whether
one draft of a figure is leaner than another, and unreliable as an absolute
number.

### And a larger caveat: the principle itself is contested

The three problems above are measurement error. There's a fourth one. Maximising
the data-ink ratio isn't straightforwardly good, and the experimental evidence
has said so for thirty years.

Gillan and Richman (Human Factors, 1994) found that higher data-ink did make
readers faster and more accurate, but concluded that the principle as stated is
too simple: non-data ink isn't one thing. An axis helps; a decorative
background doesn't; and which is which depends on the task and the graph type.
Inbar, Tractinsky and Meyer (2007) found that readers preferred graphs that were
*not* minimalist, accepting moderate reduction and rejecting Tufte's version of
it. Bateman and colleagues, in "Useful Junk?" (CHI 2010), found that embellished
charts were recalled better over the long run than plain ones. More recently the
accessibility argument has been pressed hard: a hairline on a white background
at low contrast is lean and also unreadable for a good number of people.

I haven't resolved any of that and this package doesn't try to. So I've built
the numbers here as diagnostics. "This figure spends ninety-five percent of its
ink on furniture" is a useful thing to know. "Therefore push the ratio to one"
doesn't follow. Take it to the limit and you'll produce figures that are
elegant, economical and worse to read. The audit reports, and you decide.

That's also why `theme_tufte()` takes a `grid` argument instead of refusing to
draw one. A faint grid is the honest choice whenever readers have to recover
values instead of comparing shapes.

## The lie factor

The lie factor is the size of the effect shown in the graphic divided by the
size of the effect in the data. A truthful graphic sits at one, and Tufte treats
anything outside roughly 0.95 to 1.05 as distortion.

Given the numbers directly, it reproduces his own examples. The fuel-economy
graphic in *The Visual Display* showed an eighteen percent change as a line
growing by seven hundred and eighty-three percent.

```{r lie-numeric}
lie_factor(c(18.0, 27.5), c(0.6, 5.3))
```

Given a plot, it computes the distortion a non-zero baseline introduces. That's
by far the most common way a published figure lies. The bar's length stops being
the quantity it stands for.

```{r lie-plot, fig.height = 3}
means <- data.frame(
  condition = c("Control", "Treated"),
  outcome = c(100, 110)
)

honest <- ggplot(means, aes(condition, outcome)) +
  geom_col_tufte(fill = "grey72") +
  labs(x = NULL, y = "Outcome") +
  theme_tufte()

truncated <- honest + coord_cartesian(ylim = c(95, 115))

c(honest = lie_factor(honest), truncated = lie_factor(truncated))
```

A ten percent difference drawn as a sixteen-fold one. Here they're side by
side. The second one is the figure that gets published.

```{r lie-side, fig.height = 2.6}
print(honest + labs(title = "Baseline at zero"))
print(truncated + labs(title = "Baseline at 95"))
```

### Where it misleads

The plot method only looks at bars. A truncated axis on a line chart or a dot
plot isn't necessarily a lie, since those marks encode position and not length,
and Tufte's own advice is that you can crop them freely. So `lie_factor()`
returns `1` for a scatterplot whatever its limits. That's correct, and it's easy
to misread as a clean bill of health.

## Data density

The number of entries in the data matrix divided by the area of the data
graphic, in square inches. Tufte's complaint about most published graphics is
that they're enormous and say almost nothing. A chart carrying four numbers over
half a page would have been better as a sentence.

```{r density}
data_density(lean, width = 6.5, height = 4)
```

Entries get counted as rows drawn times distinct variables mapped to aesthetics.
The area is the panel and not the whole figure, since that's what Tufte means by
"the data graphic". Constants set outside `aes()` don't count, because they
carry no data.

```{r density-sparse}
four_numbers <- data.frame(g = letters[1:4], v = c(3, 7, 5, 9))
data_density(
  ggplot(four_numbers, aes(g, v)) + geom_col_tufte() + theme_tufte(),
  width = 6.5, height = 4
)
```

Sixteen numbers per square inch sounds respectable until you notice the whole
figure carries eight of them. The measure rewards a plot for having many mapped
variables even when two of them are the same column, so read it alongside the
redundant-encoding check in the audit.

## Checking that nothing is clipped

This isn't a Tufte quantity. I think it ruins more figures than any of them.
`ggplot2` clips long text at the device edge, silently, and a subtitle that fits
on screen at the default device size won't necessarily fit in the saved file.

`check_labels_fit()` renders at the size you intend to print and measures every
text element against the space available.

```{r fit}
wordy <- lean + labs(
  subtitle = paste(rep("A subtitle that runs on rather too long", 4),
                   collapse = " ")
)

check_labels_fit(wordy, width = 6.5, height = 4)
```

It also catches axis labels that can't sit side by side, which is the other
common failure and the hardest one to see in a preview pane.

```{r fit-axis}
long_labels <- data.frame(
  condition = c("Non-post-conflict baseline", "Post-conflict, high intensity",
                "Post-conflict, low intensity"),
  value = c(3.1, 4.8, 4.0)
)

check_labels_fit(
  ggplot(long_labels, aes(condition, value)) + geom_col_tufte() + theme_tufte(),
  width = 4, height = 3
)
```

`save_tufte()` runs this check before it writes the file, so the warning shows
up while you can still do something about it.

## Putting it together

`tufte_audit()` runs all of the above alongside the criteria Tufte states
outright, and is careful about which is which.

```{r audit-bad}
bad <- ggplot(d, aes(g, y, fill = g)) +
  geom_col(stat = "summary", fun = "mean")

suppressWarnings(tufte_audit(bad, width = 6.5, height = 4))
```

Several stated criteria not met, each named with the principle it comes from,
and separately a set of measurements with no verdict attached. The same data,
drawn properly:

```{r audit-good}
good <- ggplot(d, aes(g, y)) +
  geom_tufteboxplot() +
  geom_rangeframe(sides = "l") +
  labs(x = NULL, y = "Outcome") +
  theme_tufte() +
  label_source("Simulated data, n = 400")

tufte_audit(good, width = 6.5, height = 4)
```

## Every figure in the paper at once

Auditing one plot is useful while you're drawing it. Auditing all of them, the
evening before you submit, is when it earns its keep. The figure with the
truncated subtitle is never the one you were looking at.

```{r batch}
figures <- list(
  scatter = good,
  bars = bad,
  faint = ggplot(d, aes(x, y)) +
    geom_point(colour = "grey85") + theme_tufte() + label_source("Simulated")
)

suppressWarnings(audit_figures(figures, measure = FALSE))
```

It also takes a directory, so you can check a whole replication package in one
call if the figures were saved with `saveRDS()`.

## Why there's no score

An earlier version of this package reported a score, the share of checks passed.
It's gone, for two reasons.

The first is that it meant making up thresholds. Tufte gives a testable line for
some principles and only a direction for others, so a pass or fail on the second
kind can only come from me. The version of `data_ink_ratio()` that failed a
figure below 0.5 was asserting something Tufte never wrote, in his voice.
`tufte_principles()` now marks the difference in its `criterion` column, and the
audit grades only the principles that carry one.

```{r criteria}
p <- tufte_principles()
p[p$criterion, c("principle", "source")]
```

The second is that a score needs a weighting. Saying a figure is at seventy
percent means you've decided how many missing source notes equal one pie chart,
and Tufte doesn't offer an exchange rate. So the audit reports a count of stated
criteria not met, which is comparable across figures because every figure gets
counted against the same list, plus a set of measurements you can read against
another draft.

None of this puts the remaining criteria beyond argument. They're my reading of
what Tufte states outright, and I've tried to keep that reading visible in the
source of each check instead of burying it in a number.

`tufte_principles()` is the honest inventory. The `audited` column marks which
principles a function can reach at all, and `criterion` marks which of those
Tufte states a testable line for. The ones no function can reach are listed
anyway.

```{r principles}
p <- tufte_principles()
p[!p$audited, c("principle", "source", "implemented_by")]
```

Showing comparisons, showing causality, showing multivariate data, content
counting most of all: a package can't check any of it. That's what the figure is
for. What the audit is good at is the mechanical stuff underneath, which is real
and common and which you stop seeing after the fifth draft.
