---
title: "Measuring a figure"
description: >
  What the data-ink ratio, the lie factor and data density actually compute,
  and where each of them will mislead you.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6.5, fig.height = 4, dpi = 150, out.width = "100%",
  fig.align = "center", dev = "png"
)
library(ggplot2)
library(tufter)
```

Most of what people take from Tufte is a look. The more interesting claim in
*The Visual Display of Quantitative Information* is methodological: that
statistical graphics can be evaluated, and that he can give you the quantities
to do it with. This article works through what each of those quantities means
here, how it is computed, and where it will lead you astray if you quote it
without thinking.

All data below are simulated.

```{r data}
set.seed(20260729)
n <- 400
d <- data.frame(
  x = rnorm(n, 50, 12),
  g = sample(c("Treated", "Control"), n, replace = TRUE)
)
d$y <- 12 + 0.42 * d$x + ifelse(d$g == "Treated", 3.1, 0) + rnorm(n, 0, 5)
```

## The data-ink ratio

Tufte defines the data-ink ratio as the share of a graphic's ink devoted to the
non-redundant display of data, and asks that it be pushed towards one. That is a
definition, not a procedure: no book tells you how to count ink on a screen.

`data_ink_ratio()` estimates it empirically. The plot is rendered to a bitmap
twice, once whole and once with every data layer stripped out of the panel
grobs. Each pixel contributes in proportion to how far it sits from the
background colour, so an anti-aliased edge counts as the fraction of a pixel it
actually is. What is left after subtraction is data ink.

```{r ratio-default}
base <- ggplot(d, aes(x, y)) +
  geom_point(size = 1, alpha = 0.5) +
  labs(x = "Pre-treatment score", y = "Outcome")

data_ink_ratio(base)
```

Almost nothing. The grey panel alone is most of that ink, and it does not change
when the data change.

```{r ratio-lean}
lean <- base + geom_rangeframe() + theme_tufte()
data_ink_ratio(lean)
```

The same numbers, drawn with about seven times the share of the ink doing work.

### Where it misleads

Three things are worth knowing before you quote the number.

**Overlap is counted once.** A dense scatterplot draws many points on top of
each other, and the measurement sees one blob. So dense plots understate their
own data-ink, which is the opposite of the direction you would want the bias to
run.

```{r overlap, fig.height = 3}
dense <- data.frame(x = rnorm(20000), y = rnorm(20000))
sparse <- dense[1:60, ]

lean_theme <- list(geom_rangeframe(), theme_tufte())

c(
  dense = data_ink_ratio(
    ggplot(dense, aes(x, y)) + geom_point(size = 0.4, alpha = 0.2) + lean_theme
  )$ratio,
  sparse = data_ink_ratio(
    ggplot(sparse, aes(x, y)) + geom_point(size = 0.4) + lean_theme
  )$ratio
)
```

**Redundant data-ink still counts as data-ink.** Tufte would subtract ink that
repeats information the reader already has. No measurement can tell whether a
mark is repeating something, so this estimate does not try.

**The number depends on the size you render at.** Non-data ink is largely fixed
furniture, so shrinking the canvas raises the ratio. Compare figures at the size
you intend to print them, and compare like with like.

```{r size-dependence}
vapply(
  c(3, 6.5, 12),
  function(w) data_ink_ratio(lean, width = w, height = w * 0.6)$ratio,
  numeric(1)
)
```

Treat it as a comparative instrument. It is reliable for judging whether one
version of a figure is leaner than another, and unreliable as an absolute score.

### And a larger caveat: the principle itself is contested

The three problems above are measurement error. There is a fourth problem, which
is that maximising the data-ink ratio is not straightforwardly good, and the
experimental evidence has said so for thirty years.

Gillan and Richman (Human Factors, 1994) found that higher data-ink did make
readers faster and more accurate, but concluded that the principle as stated is
too simple: non-data ink is not one thing. An axis helps; a decorative
background does not; and which is which depends on the task and the graph type.
Inbar, Tractinsky and Meyer (2007) found that readers preferred graphs that were
*not* minimalist, accepting moderate reduction and rejecting Tufte's version of
it. Bateman and colleagues, in "Useful Junk?" (CHI 2010), found that embellished
charts were recalled better over the long run than plain ones. More recently the
accessibility argument has been pressed hard: a hairline on a white background
at low contrast is lean and also unreadable for a good number of people.

I have not resolved any of that, and this package does not try to. The design
consequence is that the numbers here are meant as *descriptive diagnostics*, not
as an objective function. "This figure spends ninety-five percent of its ink on
furniture" is a useful thing to know. "Therefore push the ratio to one" does not
follow, and if you take it to the limit you will produce figures that are
elegant, economical, and worse to read. The audit reports; you decide.

This is also why `theme_tufte()` takes a `grid` argument rather than refusing to
draw one. A faint grid is the honest choice whenever readers have to recover
values rather than compare shapes.

## The lie factor

The lie factor is the size of the effect shown in the graphic divided by the
size of the effect in the data. A truthful graphic has a lie factor of one, and
Tufte treats anything outside roughly 0.95 to 1.05 as distortion.

Given the numbers directly, it reproduces his own examples. The fuel-economy
graphic in *The Visual Display* showed an eighteen percent change as a line
growing by seven hundred and eighty-three percent.

```{r lie-numeric}
lie_factor(c(18.0, 27.5), c(0.6, 5.3))
```

Given a plot, it computes the distortion introduced by a baseline that is not
zero, which is by far the most common way a real published figure lies. A bar's
length stops being the quantity it stands for.

```{r lie-plot, fig.height = 3}
means <- data.frame(
  condition = c("Control", "Treated"),
  outcome = c(100, 110)
)

honest <- ggplot(means, aes(condition, outcome)) +
  geom_col_tufte(fill = "grey72") +
  labs(x = NULL, y = "Outcome") +
  theme_tufte()

truncated <- honest + coord_cartesian(ylim = c(95, 115))

c(honest = lie_factor(honest), truncated = lie_factor(truncated))
```

A ten percent difference drawn as a sixteen-fold one. Here they are side by
side; the second is the figure that gets published.

```{r lie-side, fig.height = 2.6}
print(honest + labs(title = "Baseline at zero"))
print(truncated + labs(title = "Baseline at 95"))
```

### Where it misleads

The plot method only looks at bars. A truncated axis on a line chart or a dot
plot is not necessarily a lie, because those marks encode position rather than
length, and Tufte's own advice is that they may be cropped freely. So
`lie_factor()` returns `1` for a scatterplot regardless of its limits. That is
correct behaviour and easy to misread as a clean bill of health.

## Data density

The number of entries in the data matrix divided by the area of the data
graphic, in square inches. Tufte's complaint about most published statistical
graphics is that they are enormous and say almost nothing: a chart carrying four
numbers over half a page would have been better as a sentence.

```{r density}
data_density(lean, width = 6.5, height = 4)
```

The entries are counted as rows drawn times distinct variables mapped to
aesthetics, and the area is the panel rather than the whole figure, since that
is what Tufte means by "the data graphic". Constants set outside `aes()` do not
count, because they carry no data.

```{r density-sparse}
four_numbers <- data.frame(g = letters[1:4], v = c(3, 7, 5, 9))
data_density(
  ggplot(four_numbers, aes(g, v)) + geom_col_tufte() + theme_tufte(),
  width = 6.5, height = 4
)
```

Sixteen numbers per square inch sounds respectable until you notice the whole
figure carries eight of them. The density measure rewards a plot for having many
mapped variables even when those variables are the same column twice, so read it
alongside the redundant-encoding check in the audit.

## Checking that nothing is clipped

Not a Tufte quantity, but the failure that ruins more figures than any of them.
`ggplot2` does not wrap long text: it clips it at the device edge, silently, and
a subtitle that fits on screen at the default device size is not a subtitle that
fits in the saved file.

`check_labels_fit()` renders at the size you intend to print and measures every
text element against the space available.

```{r fit}
wordy <- lean + labs(
  subtitle = paste(rep("A subtitle that runs on rather too long", 4),
                   collapse = " ")
)

check_labels_fit(wordy, width = 6.5, height = 4)
```

It also catches axis labels that cannot sit side by side, which is the other
common failure and the one that is hardest to see in a preview pane.

```{r fit-axis}
long_labels <- data.frame(
  condition = c("Non-post-conflict baseline", "Post-conflict, high intensity",
                "Post-conflict, low intensity"),
  value = c(3.1, 4.8, 4.0)
)

check_labels_fit(
  ggplot(long_labels, aes(condition, value)) + geom_col_tufte() + theme_tufte(),
  width = 4, height = 3
)
```

`save_tufte()` runs this check before it writes the file, so the warning arrives
while you can still act on it.

## Putting it together

`tufte_audit()` runs all of the above plus the structural checks, and scores
them.

```{r audit-bad}
bad <- ggplot(d, aes(g, y, fill = g)) +
  geom_col(stat = "summary", fun = "mean")

suppressWarnings(tufte_audit(bad, width = 6.5, height = 4))
```

Six separate problems, named: the tinted panel, the minor gridlines, a legend
that should have been a direct label, the grouping variable encoded twice, no
source note, and four numbers spread over seventeen square inches. The same
data, drawn properly:

```{r audit-good}
good <- ggplot(d, aes(g, y)) +
  geom_tufteboxplot() +
  geom_rangeframe(sides = "l") +
  labs(x = NULL, y = "Outcome") +
  theme_tufte() +
  label_source("Simulated data, n = 400")

tufte_audit(good, width = 6.5, height = 4)
```

## Every figure in the paper at once

Auditing one plot is useful while you are drawing it. Auditing all of them, the
evening before you submit, is when it earns its keep, because the figure with
the truncated subtitle is never the one you were looking at.

```{r batch}
figures <- list(
  scatter = good,
  bars = bad,
  faint = ggplot(d, aes(x, y)) +
    geom_point(colour = "grey85") + theme_tufte() + label_source("Simulated")
)

suppressWarnings(audit_figures(figures, measure = FALSE))
```

It also takes a directory, so a replication package whose figures were saved
with `saveRDS()` can be checked in one call.

## What the score is not

The score is the share of applicable checks passed, and it is a prompt rather
than a verdict. A figure can pass every check and still be pointless, because
Tufte's first principle is that content counts most of all, and no function
evaluates that.

`tufte_principles()` is the honest inventory. Its `audited` column marks which
principles a function can verify and which it cannot, and the ones it cannot are
listed anyway.

```{r principles}
p <- tufte_principles()
p[!p$audited, c("principle", "source", "implemented_by")]
```

Showing comparisons, showing causality, showing multivariate data, and content
counting most of all are not things a package can check. They are what the
figure is for. What the audit is good for is the mechanical failures underneath
them: the ones that are real, that are common, and that authors stop seeing
after the fifth draft.
