---
title: "Measuring a figure"
description: >
  How the graphical measurements are estimated and what to check when
  interpreting them.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6.5, fig.height = 4, dpi = 300, out.width = "100%",
  fig.align = "center", dev = "png"
)
library(ggplot2)
library(tufter)
```

What can a number tell us about a figure? Tufte proposes several measures
in *The Visual Display of Quantitative Information*. I built these functions
to make those measures easier to examine.

Here I explain how they're estimated and what their limits mean in practice.
All the data below are simulated.

```{r data}
set.seed(20260729)
n <- 400
d <- data.frame(
  x = rnorm(n, 50, 12),
  g = sample(c("Treated", "Control"), n, replace = TRUE)
)
d$y <- 12 + 0.42 * d$x + ifelse(d$g == "Treated", 3.1, 0) + rnorm(n, 0, 5)
```

## The data-ink ratio

Tufte defines data-ink ratio as the share of ink devoted to the non-redundant
display of data. `data_ink_ratio()` estimates it by rendering a plot twice,
with and without its data layers.

Each pixel is weighted by its distance from the background colour. The
difference between the two renderings estimates the data ink. This is an
approximation, including at anti-aliased edges.

```{r ratio-default}
base <- ggplot(d, aes(x, y)) +
  geom_point(size = 1, alpha = 0.5) +
  labs(x = "Pre-treatment score", y = "Outcome")

r_base <- data_ink_ratio(base)
r_base
```

Much of the estimated ink here comes from the panel background and grid.
Those elements are counted as non-data ink.

```{r ratio-lean}
lean <- base + geom_rangeframe() + theme_tufte()
r_lean <- data_ink_ratio(lean)
r_lean
```

The data are the same. The estimated share of data ink is
`r sprintf("%.1f", r_lean$ratio / r_base$ratio)` times the original share.

### Where it misleads

Overlapping marks count once. A dense group of points can therefore produce
less measured data ink than the same points spread apart.

```{r overlap, fig.height = 3}
dense <- data.frame(x = rnorm(20000), y = rnorm(20000))
sparse <- dense[1:60, ]

lean_theme <- list(geom_rangeframe(), theme_tufte())

c(
  dense = data_ink_ratio(
    ggplot(dense, aes(x, y)) + geom_point(size = 0.4, alpha = 0.2) + lean_theme
  )$ratio,
  sparse = data_ink_ratio(
    ggplot(sparse, aes(x, y)) + geom_point(size = 0.4) + lean_theme
  )$ratio
)
```

Redundant marks still count as data ink. The function can't determine
whether they repeat information, so large filled shapes can produce high
ratios even when a smaller mark would show the same value.

The estimate also depends on the rendering size. Compare drafts at the same
width, height and resolution.

```{r size-dependence}
vapply(
  c(3, 6.5, 12),
  function(w) data_ink_ratio(lean, width = w, height = w * 0.6)$ratio,
  numeric(1)
)
```

I use the ratio to compare drafts at the same dimensions. It doesn't give
me a target to optimize.

### And a larger caveat: the principle itself is contested

The measurement's limits are one issue. Whether a higher ratio helps readers
is a separate empirical question.

Gillan and Richman (Human Factors, 1994) found that higher data-ink did make
readers faster and more accurate, but concluded that the principle as stated is
too simple: non-data ink isn't one thing. An axis helps; a decorative
background doesn't; and which is which depends on the task and the graph type.
Inbar, Tractinsky and Meyer (2007) found that readers preferred graphs that were
*not* minimalist, accepting moderate reduction and rejecting Tufte's version of
it. Bateman and colleagues, in "Useful Junk?" (CHI 2010), found that embellished
charts were recalled better over the long run than plain ones. More recently the
accessibility argument has been pressed hard: a hairline on a white background
at low contrast is lean and also unreadable for a good number of people.

I treat the estimate as a diagnostic. Knowing how much ink comes from data
layers can be useful, but it doesn't establish that removing more ink will
help. A grid, for example, can make values easier to estimate. That's why
`theme_tufte()` lets you keep one.

## The lie factor

The lie factor compares the proportional change shown in the graphic with
the proportional change in the data. A value of one means they agree.
Tufte's reference band is roughly 0.95 to 1.05.

The numeric method can be used with examples such as the fuel-economy
graphic in *The Visual Display*.

```{r lie-numeric}
lie_factor(c(18.0, 27.5), c(0.6, 5.3))
```

For a supported bar chart, the plot method measures the distortion caused
by a non-zero baseline. Bar lengths then show different proportions from
the underlying values.

```{r lie-plot, fig.height = 3}
means <- data.frame(
  condition = c("Control", "Treated"),
  outcome = c(100, 110)
)

honest <- ggplot(means, aes(condition, outcome)) +
  geom_col_tufte(fill = "grey72") +
  labs(x = NULL, y = "Outcome") +
  theme_tufte()

truncated <- honest + coord_cartesian(ylim = c(95, 115))

c(honest = lie_factor(honest), truncated = lie_factor(truncated))
```

The cropped version makes the relative difference look much larger. Here
are the two versions side by side.

```{r lie-side, fig.height = 2.6}
print(honest + labs(title = "Baseline at zero"))
print(truncated + labs(title = "Baseline at 95"))
```

### Where it misleads

The plot method checks bars. A line or dot plot can reasonably use an axis
that excludes zero because its marks show position. `lie_factor()` returns
`1` for a plot without bars, but that doesn't establish that the figure is
accurate in other respects.

## Data density

Data density estimates entries per square inch of graphic. I use it to
compare how much space different drafts give to the same information.

```{r density}
data_density(lean, width = 6.5, height = 4)
```

The function multiplies pooled rows by the number of distinct mapped
variables. Constants outside `aes()` don't count. Panel area is estimated
from the rendered plot; if that estimate isn't available, the whole canvas
is used.

```{r density-sparse}
four_numbers <- data.frame(g = letters[1:4], v = c(3, 7, 5, 9))
data_density(
  ggplot(four_numbers, aes(g, v)) + geom_col_tufte() + theme_tufte(),
  width = 6.5, height = 4
)
```

Read density alongside the number of entries. A compact plot can have a
high density while showing only a few numbers. Layers that combine data
sources or statistical summaries also need manual interpretation.

## Checking that nothing is clipped

A label that fits in a preview can be cut off at the size you export.
`check_labels_fit()` estimates the space needed by titles, axes, legends and
facet strips at the requested dimensions. It doesn't measure text drawn
inside the panel.

```{r fit}
wordy <- lean + labs(
  subtitle = paste(rep("A subtitle that runs on rather too long", 4),
                   collapse = " ")
)

check_labels_fit(wordy, width = 6.5, height = 4)
```

The check also looks for crowded axis labels. I'd still inspect the
exported file, since the function can't detect every overlap.

```{r fit-axis}
long_labels <- data.frame(
  condition = c("Non-post-conflict baseline", "Post-conflict, high intensity",
                "Post-conflict, low intensity"),
  value = c(3.1, 4.8, 4.0)
)

check_labels_fit(
  ggplot(long_labels, aes(condition, value)) + geom_col_tufte() + theme_tufte(),
  width = 4, height = 3
)
```

`save_tufte()` runs the check before writing the file. Use `strict = TRUE`
if a failed or unavailable check should stop the save.

## Putting it together

`tufte_audit()` runs these checks and reports the measurements together.

```{r audit-bad}
bad <- ggplot(d, aes(g, y, fill = g)) +
  geom_col(stat = "summary", fun = "mean")

suppressWarnings(tufte_audit(bad, width = 6.5, height = 4))
```

The output identifies unmet criteria and reports measurements separately.
Here's another version of the same data with the suggested changes.

```{r audit-good}
good <- ggplot(d, aes(g, y)) +
  geom_tufteboxplot() +
  geom_rangeframe(sides = "l") +
  labs(x = NULL, y = "Outcome") +
  theme_tufte() +
  label_source("Simulated data, n = 400")

tufte_audit(good, width = 6.5, height = 4)
```

## Every figure in the paper at once

I use `audit_figures()` to review a set of figures together. It lists the
figures with the most unmet criteria first, along with any checks that
couldn't run.

```{r batch}
figures <- list(
  scatter = good,
  bars = bad,
  faint = ggplot(d, aes(x, y)) +
    geom_point(colour = "grey85") + theme_tufte() + label_source("Simulated")
)

suppressWarnings(audit_figures(figures, measure = FALSE))
```

You can also pass a directory containing plots saved with `saveRDS()`.

## Why there's no score

A pass applies to the check described. It doesn't certify the whole figure.
Label fitting leaves panel text for manual review. Contrast checks don't
resolve every overlap or filled label background. Unsupported lie-factor
comparisons can return an unavailable result. The batch table's `skipped`
column identifies checks that couldn't run.

I've kept measurements without a stated threshold separate from graded
criteria. The `criterion` column in `tufte_principles()` records that choice.

```{r criteria}
p <- tufte_principles()
p[p$criterion, c("principle", "source")]
```

Combining everything into a percentage would require a judgement about
how much each criterion matters. I haven't made that judgement for you.
The audit gives counts and measurements you can examine individually.

These checks reflect my reading of Tufte's principles. You can inspect the
source of each check and disagree with how I've implemented it.
`tufte_principles()` also records which principles the package can't check.

```{r principles}
p <- tufte_principles()
p[!p$audited, c("principle", "source", "implemented_by")]
```

I still need to ask whether the figure answers the research question and
whether the design supports my interpretation. The measurements help me
inspect the drawing; they don't answer those questions.
