---
title: "Examples with simulated data"
description: >
  Plotting examples using simulated survey and experimental data, with code
  you can run yourself.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6.5, fig.height = 4, dpi = 300, out.width = "100%",
  fig.align = "center", dev = "png"
)
library(ggplot2)
library(tufter)
```

Here are examples using simulated data. I've used settings familiar from
survey experiments and discrimination research, including callback rates and
treatment effects across conditions. The numbers are illustrative.

Run the setup and simulation helper before the examples. Each section sets
its seed so you can reproduce the figures.

## Simulating the data

This helper creates respondents, randomly assigns a condition, and generates
an outcome with a modest treatment effect.

```{r sim-helper}
simulate_experiment <- function(n = 900,
                                conditions = c("Control", "Information",
                                               "Norms", "Both"),
                                effects = c(0, 0.28, 0.19, 0.41),
                                seed = 20260729) {
  set.seed(seed)
  condition <- factor(sample(conditions, n, replace = TRUE),
                      levels = conditions)
  age <- round(rnorm(n, 46, 15))
  age <- pmin(pmax(age, 18), 85)
  effect <- effects[match(condition, conditions)]

  data.frame(
    id = seq_len(n),
    condition = condition,
    age = age,
    # Outcome on a 1 to 7 scale, with a mild age gradient.
    support = pmin(pmax(
      rnorm(n, 4.1 + effect + 0.012 * (age - 46), 1.15), 1
    ), 7)
  )
}

experiment <- simulate_experiment()
str(experiment)
```

## Distributions: the box plot with the box erased

`geom_tufteboxplot()` shows the distribution with whisker lines and a median
mark. `geom_rangeframe(sides = "l")` adds a vertical axis line spanning the
data.

```{r boxplot}
ggplot(experiment, aes(condition, support)) +
  geom_tufteboxplot() +
  geom_rangeframe(sides = "l") +
  labs(
    x = NULL, y = "Support (1-7)",
    title = "Support for the policy, by experimental condition"
  ) +
  theme_tufte() +
  label_source("Simulated data, n = 900")
```

The default, `"point"`, leaves a gap for the interquartile range and places
a dot at the median. `"line"` uses a thicker interquartile segment.
`"offset"` places that segment beside the whiskers. I'd compare them at the
size the figure will be printed.

```{r boxplot-variants, fig.height = 3}
for (variant in c("line", "offset")) {
  print(
    ggplot(experiment, aes(condition, support)) +
      geom_tufteboxplot(type = variant) +
      geom_rangeframe(sides = "l") +
      labs(x = NULL, y = "Support", title = paste0('type = "', variant, '"')) +
      theme_tufte()
  )
}
```

## Range frames and quartile frames

A range frame spans the observed values. A quartile frame also marks the
five-number summary, with labels supplied by `quartile_breaks()`.

All five breaks are kept by default. If labels crowd together, use `min_gap`
to omit some of them. Run `check_labels_fit()` and inspect the saved figure.

```{r frames}
ggplot(experiment, aes(age, support)) +
  geom_point(alpha = 0.25, size = 1) +
  geom_quartileframe() +
  scale_x_continuous(breaks = quartile_breaks(experiment$age)) +
  scale_y_continuous(breaks = quartile_breaks(experiment$support)) +
  labs(x = "Age", y = "Support") +
  theme_tufte()
```

`geom_dotdash()` adds a short tick for every observation along each margin.
Turn off ordinary theme ticks if you'd like these marks to replace them.

```{r dotdash}
ggplot(experiment, aes(age, support)) +
  geom_point(alpha = 0.25, size = 1) +
  geom_dotdash() +
  labs(x = "Age", y = "Support") +
  theme_tufte(ticks = FALSE)
```

## Bars with the gridlines erased through them

This bar design uses gaps through the bars as gridlines. They help readers
estimate values without adding a separate grid across the panel.

The example shows simulated callback rates by applicant name and occupation.

```{r bars}
set.seed(11)
callbacks <- data.frame(
  name = rep(c("Emily", "Greg", "Lakisha", "Jamal"), each = 2),
  occupation = rep(c("Sales", "Administrative"), 4)
)
callbacks$rate <- round(
  c(10.8, 9.4, 10.2, 9.1, 6.6, 5.9, 6.4, 5.5) + rnorm(8, 0, 0.3), 1
)

ggplot(callbacks, aes(name, rate)) +
  geom_col_tufte(fill = "grey72") +
  facet_tufte(~ occupation) +
  labs(x = NULL, y = "Callback rate (%)",
       title = "Simulated callback rates by name and occupation") +
  theme_tufte() +
  label_source("Simulated data; not real audit-study results")
```

The bars start at zero. `tufte_audit()` checks the baseline, and
`lie_factor()` estimates distortion in supported bar comparisons.

## Dots, when a zero baseline isn't wanted

A dot plot is useful when the values occupy a narrow range far from zero.
The dots show position on the scale. Leader lines help readers follow each
row to its value.

I've sorted the categories by value so rank is easy to see.

```{r cleveland, fig.height = 3}
support <- aggregate(support ~ condition, experiment, mean)

ggplot(support, aes(support, stats::reorder(condition, support))) +
  geom_cleveland_dot() +
  labs(x = "Mean support (1-7)", y = NULL) +
  theme_tufte() +
  label_source("Simulated data, n = 900")
```

## Banking the aspect ratio

The same series looks steeper in a taller panel. `bank_to_45()` uses
Cleveland's banking approach to suggest a height that brings the median
absolute slope near 45 degrees.

```{r banking}
cycles <- data.frame(
  month = 1:240,
  value = sin(seq(0, 12 * pi, length.out = 240)) + seq(0, 2, length.out = 240)
)

series <- ggplot(cycles, aes(month, value)) +
  geom_line(linewidth = 0.3) +
  geom_rangeframe(sides = "l") +
  labs(x = "Month", y = "Index") +
  theme_tufte()

banked <- bank_to_45(series, width = 6.5)
banked
```

At this height, I find the gradual upward trend easy to see. The individual
cycles are compressed.

```{r banked-figure, fig.height = 1.4}
series
```

Here's a taller version. It gives the cycles more space, which may be
useful if their pattern is what you want to compare.

```{r unbanked-figure, fig.height = 4}
series
```

The choice depends on the question. I'd use the suggested height as a
starting point and compare it with another version.

## Is it dark enough to read?

`check_contrast()` compares mark and text colours with the background. Its
default minima are 4.5 to 1 for text and 3 to 1 for marks. It doesn't resolve
every overlapping mark or separately filled background.

```{r contrast}
check_contrast(
  ggplot(experiment, aes(age, support)) +
    geom_point(colour = "grey80", size = 0.8) +
    theme_tufte()
)
```

Grey 80 has low contrast against white. I'd use a darker colour for marks
that readers need to distinguish individually.

## Small multiples

Repeating the same plot for each condition can make comparisons easier.
`facet_tufte()` fixes the scales and warns if you change them.

```{r multiples, fig.height = 3.2}
ggplot(experiment, aes(age, support)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, colour = "grey20", linewidth = 0.4,
              formula = y ~ x) +
  geom_rangeframe() +
  facet_tufte(~ condition, ncol = 4) +
  labs(x = "Age", y = "Support") +
  theme_tufte()
```

## Direct labelling instead of a legend

Put the series name at the end of the line with `geom_text_last()`. Readers
can then identify a series where they're looking at its values.

```{r direct}
set.seed(4)
countries <- c("Australia", "Japan", "Korea", "New Zealand")
panel <- do.call(rbind, lapply(seq_along(countries), function(i) {
  data.frame(
    year = 2000:2024,
    country = countries[i],
    trust = cumsum(rnorm(25, 0.1 * (i - 2.5), 1.1)) + 45 + 4 * i
  )
}))

ggplot(panel, aes(year, trust, colour = country)) +
  geom_line(linewidth = 0.4) +
  # Labels take a fixed dark grey rather than the line colour: a label in the
  # lightest grey of a sequential palette is legible as a line and not as text.
  geom_text_last(aes(label = country, group = country), size = 3, colour = "grey15") +
  geom_rangeframe(sides = "l") +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.18))) +
  scale_colour_tufte("grey") +
  labs(x = NULL, y = "Institutional trust index") +
  theme_tufte() +
  theme(legend.position = "none") +
  label_source("Simulated panel, four countries, 2000-2024")
```

These labels can overlap because `geom_text_last()` doesn't move them
apart. Try `nudge_y` or `ggrepel::geom_text_repel()` if the endpoints are close.
The slopegraph function below includes label spacing, though dense plots
still need inspection.

## Slopegraphs

A slopegraph compares two periods. The labels give the values and line
crossings show changes in rank. I find it useful when both the levels and
the changes matter.

```{r slopegraph, fig.height = 5.5}
set.seed(7)
units <- c("Victoria", "New South Wales", "Queensland", "South Australia",
           "Western Australia", "Tasmania")
before <- round(runif(6, 22, 41), 1)
change <- round(rnorm(6, 3.5, 5), 1)

wave <- data.frame(
  state = rep(units, each = 2),
  wave = rep(c("2015", "2025"), 6),
  value = as.vector(rbind(before, before + change))
)

slopegraph(wave, wave, value, state) +
  scale_x_continuous(breaks = 1:2, labels = c("2015", "2025"),
                     position = "top", expand = expansion(mult = 0.5)) +
  labs(
    title = "Share reporting contact with a migrant neighbour",
    subtitle = "Simulated two-wave panel, percentage points"
  )
```

Set `direction_colour = TRUE` to distinguish increases from decreases by
colour. I'd use that when direction is the comparison I want to emphasize.

```{r slopegraph-colour, fig.height = 5.5}
slopegraph(wave, wave, value, state, direction_colour = TRUE) +
  scale_x_continuous(breaks = 1:2, labels = c("2015", "2025"),
                     position = "top", expand = expansion(mult = 0.5))
```

## Sparklines

Sparklines fit a series into a small space. Here the band shows its
interquartile range and the dots identify the extremes. Each series has its
own vertical scale, so compare patterns rather than heights across rows.

```{r sparklines, fig.height = 3}
set.seed(21)
indicators <- do.call(rbind, lapply(
  c("Turnout", "Trust in parliament", "Union density", "Newspaper readership"),
  function(nm) {
    data.frame(
      quarter = 1:80,
      series = nm,
      value = cumsum(rnorm(80, 0, 1)) + 50
    )
  }
))

sparklines(indicators, quarter, value, series)
```

Use `sparkline()` for one series. `sparkline_grob()` returns a grob you can
place in a table or another graphic.

```{r sparkline-one, fig.width = 2.4, fig.height = 0.5, out.width = "240px"}
set.seed(3)
sparkline(cumsum(rnorm(60)) + 20)
```

## Colour as a code

The palettes serve different uses. `"grey"` gives an ordered sequence.
`"accent"` adds one red series to a set of greys. `"muted"` uses earth tones,
and `"divergent"` provides a blue-to-red scale with a neutral midpoint.
Check their contrast in the figure where you'll use them.

```{r palettes, fig.height = 2.4}
pals <- c("grey", "accent", "muted", "divergent")
swatches <- do.call(rbind, lapply(pals, function(p) {
  cols <- tufte_pal(p)(5)
  data.frame(palette = p, i = seq_along(cols), colour = cols)
}))

ggplot(swatches, aes(i, palette, fill = colour)) +
  geom_tile(colour = "white", linewidth = 2) +
  scale_fill_identity() +
  labs(x = NULL, y = NULL) +
  theme_tufte(ticks = FALSE) +
  theme(axis.text.x = element_blank())
```

Here I've used grey for the comparison series and an accent for the one I
want readers to notice.

```{r accent}
focus <- panel
focus$highlight <- ifelse(focus$country == "Korea", "Korea", "Other")

ggplot(focus, aes(year, trust, group = country, colour = highlight)) +
  geom_line(linewidth = 0.4) +
  geom_rangeframe(sides = "l") +
  scale_colour_manual(values = c(Korea = "#c8102e", Other = "#bfbfbf")) +
  geom_text_last(
    data = subset(focus, country == "Korea"),
    aes(label = country), size = 3, colour = "#c8102e"
  ) +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.1))) +
  labs(x = NULL, y = "Institutional trust index") +
  theme_tufte() +
  theme(legend.position = "none")
```

## Auditing the result

`tufte_audit()` checks the available criteria and reports measurements that
you can compare across drafts.

```{r audit}
final <- ggplot(experiment, aes(condition, support)) +
  geom_tufteboxplot() +
  geom_rangeframe(sides = "l") +
  labs(x = NULL, y = "Support (1-7)") +
  theme_tufte() +
  label_source("Simulated data, n = 900")

tufte_audit(final, width = 6.5, height = 4)
```

The [measuring article](measuring.html) explains how the estimates work and
where they need careful interpretation.
