---
title: "Examples with simulated data"
description: >
  Every graphical form in the package, drawn from simulated data you can
  regenerate on your own machine.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 6.5, fig.height = 4, dpi = 150, out.width = "100%",
  fig.align = "center", dev = "png"
)
library(ggplot2)
library(tufter)
```

Everything below runs on simulated data. No file is read, nothing is
downloaded, and every chunk is reproducible from the seed at the top of its
section, so you can paste any of them into a fresh session and get the same
figure. The scenarios are the ones I actually meet in survey-experimental work:
treatment effects across conditions, audit-study callback rates, cross-national
indicators over time, and before-and-after comparisons.

## Simulating the data

One helper, used throughout. It generates a small survey experiment: a set of
respondents, a randomly assigned condition, and an outcome with a real but
modest treatment effect.

```{r sim-helper}
simulate_experiment <- function(n = 900,
                                conditions = c("Control", "Information",
                                               "Norms", "Both"),
                                effects = c(0, 0.28, 0.19, 0.41),
                                seed = 20260729) {
  set.seed(seed)
  condition <- factor(sample(conditions, n, replace = TRUE),
                      levels = conditions)
  age <- round(rnorm(n, 46, 15))
  age <- pmin(pmax(age, 18), 85)
  effect <- effects[match(condition, conditions)]

  data.frame(
    id = seq_len(n),
    condition = condition,
    age = age,
    # Outcome on a 1 to 7 scale, with a mild age gradient.
    support = pmin(pmax(
      rnorm(n, 4.1 + effect + 0.012 * (age - 46), 1.15), 1
    ), 7)
  )
}

experiment <- simulate_experiment()
str(experiment)
```

## Distributions: the box plot with the box erased

The box in a box plot is a container for four numbers that a line and a dot can
hold on their own. `geom_tufteboxplot()` erases it, keeping between a third and
a fifth of the ink. `geom_rangeframe(sides = "l")` puts an axis line only where
the data are.

```{r boxplot}
ggplot(experiment, aes(condition, support)) +
  geom_tufteboxplot() +
  geom_rangeframe(sides = "l") +
  labs(
    x = NULL, y = "Support (1-7)",
    title = "Support for the policy, by experimental condition"
  ) +
  theme_tufte() +
  label_source("Simulated data, n = 900")
```

The three variants trade ink for legibility. `"point"` is the default and the
sparest: whiskers with a gap, and a dot at the median. `"line"` keeps a thick
interquartile segment. `"offset"` shifts that segment to one side, which is what
you want when the whiskers are short.

```{r boxplot-variants, fig.height = 3}
for (variant in c("line", "offset")) {
  print(
    ggplot(experiment, aes(condition, support)) +
      geom_tufteboxplot(type = variant) +
      geom_rangeframe(sides = "l") +
      labs(x = NULL, y = "Support", title = paste0('type = "', variant, '"')) +
      theme_tufte()
  )
}
```

## Range frames and quartile frames

A panel border is the same box whatever the numbers are. A range frame spans
only the data, so the frame reports the minimum and maximum for free. A
quartile frame breaks that line at the quartiles, so the axis carries the whole
five-number summary; `quartile_breaks()` puts the printed labels in the same
places, and drops any that would collide.

```{r frames}
ggplot(experiment, aes(age, support)) +
  geom_point(alpha = 0.25, size = 1) +
  geom_quartileframe() +
  scale_x_continuous(breaks = quartile_breaks(experiment$age)) +
  scale_y_continuous(breaks = quartile_breaks(experiment$support)) +
  labs(x = "Age", y = "Support") +
  theme_tufte()
```

`geom_dotdash()` goes further still, replacing the frame with the data
themselves: a short tick at every observation, on both margins. Turn the theme
ticks off, or you get two sets of marks saying the same thing.

```{r dotdash}
ggplot(experiment, aes(age, support)) +
  geom_point(alpha = 0.25, size = 1) +
  geom_dotdash() +
  labs(x = "Age", y = "Support") +
  theme_tufte(ticks = FALSE)
```

## Bars with the gridlines erased through them

Bar charts need gridlines, because readers recover values from bar heights. But
a gridline crossing a bar sits on top of ink that already encodes that value, so
Tufte erases it there rather than drawing over the bar.

Here is a simulated audit study: callback rates by applicant name and
occupation.

```{r bars}
set.seed(11)
callbacks <- data.frame(
  name = rep(c("Emily", "Greg", "Lakisha", "Jamal"), each = 2),
  occupation = rep(c("Sales", "Administrative"), 4)
)
callbacks$rate <- round(
  c(10.8, 9.4, 10.2, 9.1, 6.6, 5.9, 6.4, 5.5) + rnorm(8, 0, 0.3), 1
)

ggplot(callbacks, aes(name, rate)) +
  geom_col_tufte(fill = "grey72") +
  facet_tufte(~ occupation) +
  labs(x = NULL, y = "Callback rate (%)",
       title = "Simulated callback rates by name and occupation") +
  theme_tufte() +
  label_source("Simulated data; not real audit-study results")
```

Note the bars start at zero. `tufte_audit()` will tell you if they do not, and
`lie_factor()` will tell you by how much the figure exaggerates.

## Dots, when a zero baseline is not wanted

Sometimes zero is a long way from the data and starting there wastes most of the
panel. That is the case for using dots rather than bars: a dot encodes its value
by position, so it can be read against a scale that excludes zero without
lying about proportions, and it costs a fraction of the ink. Cleveland's leader
lines let the eye run along a row without drifting into the next one.

Sort before plotting. An alphabetical dot plot throws away the form's main
advantage, which is that rank is visible at a glance.

```{r cleveland, fig.height = 3}
support <- aggregate(support ~ condition, experiment, mean)

ggplot(support, aes(support, stats::reorder(condition, support))) +
  geom_cleveland_dot() +
  labs(x = "Mean support (1-7)", y = NULL) +
  theme_tufte() +
  label_source("Simulated data, n = 900")
```

## Banking the aspect ratio

The same series looks like a gentle drift or a cliff depending only on how tall
the panel is, and neither reading is the data's fault. Cleveland's rule is that
slope is judged most accurately near 45 degrees, and the aspect ratio is what
puts it there. `bank_to_45()` computes the height that does it.

```{r banking}
cycles <- data.frame(
  month = 1:240,
  value = sin(seq(0, 12 * pi, length.out = 240)) + seq(0, 2, length.out = 240)
)

series <- ggplot(cycles, aes(month, value)) +
  geom_line(linewidth = 0.3) +
  geom_rangeframe(sides = "l") +
  labs(x = "Month", y = "Index") +
  theme_tufte()

banked <- bank_to_45(series, width = 6.5)
banked
```

Drawn at roughly that height, the panel is short and wide, the rising and
falling flanks of each cycle sit near 45 degrees, and the slow upward drift
underneath the oscillation is the first thing you see:

```{r banked-figure, fig.height = 1.4}
series
```

Here is the same data in a conventionally proportioned panel. Nothing is
hidden, and for reading the individual cycles it is arguably the better
picture. But the vertical stretch exaggerates every flank towards the
vertical, the oscillation dominates, and the trend it is riding on takes
noticeably longer to notice:

```{r unbanked-figure, fig.height = 4}
series
```

Which of those you want depends on the question. Banking is a rule for reading
*slopes*, so it helps when the rate of change is the finding and hurts when the
levels are. It is a defensible default and not an obligation.

## Is it dark enough to read?

Erasing ink is a virtue only up to the point where what survives can still be
seen. `check_contrast()` measures every colour the plot draws with against the
background, using the WCAG minima of 4.5 to 1 for text and 3 to 1 for marks.

```{r contrast}
check_contrast(
  ggplot(experiment, aes(age, support)) +
    geom_point(colour = "grey80", size = 0.8) +
    theme_tufte()
)
```

Grey 80 on white is elegant and, for a good number of readers, invisible.

## Small multiples

The answer to multivariate data is repetition rather than complication: the same
graphic, at the same scale, once per condition. `facet_tufte()` fixes the
scales, and warns if you try to free them, because free scales destroy the
comparison the design exists to make.

```{r multiples, fig.height = 3.2}
ggplot(experiment, aes(age, support)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, colour = "grey20", linewidth = 0.4,
              formula = y ~ x) +
  geom_rangeframe() +
  facet_tufte(~ condition, ncol = 4) +
  labs(x = "Age", y = "Support") +
  theme_tufte()
```

## Direct labelling instead of a legend

A legend makes the reader look away, hold a colour in memory, look back, and
match. `geom_text_last()` puts the name where the eye already is: at the end of
the line.

```{r direct}
set.seed(4)
countries <- c("Australia", "Japan", "Korea", "New Zealand")
panel <- do.call(rbind, lapply(seq_along(countries), function(i) {
  data.frame(
    year = 2000:2024,
    country = countries[i],
    trust = cumsum(rnorm(25, 0.1 * (i - 2.5), 1.1)) + 45 + 4 * i
  )
}))

ggplot(panel, aes(year, trust, colour = country)) +
  geom_line(linewidth = 0.4) +
  # Labels take a fixed dark grey rather than the line colour: a label in the
  # lightest grey of a sequential palette is legible as a line and not as text.
  geom_text_last(aes(label = country), size = 3, colour = "grey15") +
  geom_rangeframe(sides = "l") +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.18))) +
  scale_colour_tufte("grey") +
  labs(x = NULL, y = "Institutional trust index") +
  theme_tufte() +
  theme(legend.position = "none") +
  label_source("Simulated panel, four countries, 2000-2024")
```

`geom_text_last()` places each label at its series' final point and does no
collision avoidance, so two series ending at nearly the same value will
overprint. When that happens, nudge one with `nudge_y`, or swap in
`ggrepel::geom_text_repel()` on the same data. A slopegraph, below, solves the
problem properly by spreading colliding labels apart itself.

## Slopegraphs

A slopegraph shows before-and-after for many units at once. Each unit is a line;
the slope is the change, the height is the level, and the crossings show which
units changed rank. Every number is printed on the graphic, which makes the y
axis redundant, so it goes. The table and the figure become the same object.

```{r slopegraph, fig.height = 5.5}
set.seed(7)
units <- c("Victoria", "New South Wales", "Queensland", "South Australia",
           "Western Australia", "Tasmania")
before <- round(runif(6, 22, 41), 1)
change <- round(rnorm(6, 3.5, 5), 1)

wave <- data.frame(
  state = rep(units, each = 2),
  wave = rep(c("2015", "2025"), 6),
  value = as.vector(rbind(before, before + change))
)

slopegraph(wave, wave, value, state) +
  labs(
    title = "Share reporting contact with a migrant neighbour",
    subtitle = "Simulated two-wave panel, percentage points"
  )
```

`direction_colour = TRUE` colours the lines by whether the unit rose or fell,
which is worth it only when the direction is the finding.

```{r slopegraph-colour, fig.height = 5.5}
slopegraph(wave, wave, value, state, direction_colour = TRUE)
```

## Sparklines

A sparkline is word-sized: small enough to sit inside a sentence, with the
normal range as a grey band, dots at the extremes, and the final value printed
at the end. Each series keeps its own vertical scale, because a sparkline
reports the shape of one series rather than inviting comparison of levels.

```{r sparklines, fig.height = 3}
set.seed(21)
indicators <- do.call(rbind, lapply(
  c("Turnout", "Trust in parliament", "Union density", "Newspaper readership"),
  function(nm) {
    data.frame(
      quarter = 1:80,
      series = nm,
      value = cumsum(rnorm(80, 0, 1)) + 50
    )
  }
))

sparklines(indicators, quarter, value, series)
```

A single sparkline, at the size it is meant to be printed, is
`sparkline()`. `sparkline_grob()` returns a grob, so one can be dropped into a
table cell or an inline chunk.

```{r sparkline-one, fig.width = 2.4, fig.height = 0.5, out.width = "240px"}
set.seed(3)
sparkline(cumsum(rnorm(60)) + 20)
```

## Colour as a code

Four palettes, each answering a different question. `"grey"` encodes an ordered
variable without introducing a second, unwanted, categorical signal.
`"accent"` is greys plus one signal red, for when exactly one series matters and
the rest are context. `"muted"` is desaturated earth tones that sit behind
annotation without fighting it. `"divergent"` handles signed quantities with a
neutral rather than a white midpoint.

```{r palettes, fig.height = 2.4}
pals <- c("grey", "accent", "muted", "divergent")
swatches <- do.call(rbind, lapply(pals, function(p) {
  cols <- tufte_pal(p)(5)
  data.frame(palette = p, i = seq_along(cols), colour = cols)
}))

ggplot(swatches, aes(i, palette, fill = colour)) +
  geom_tile(colour = "white", linewidth = 2) +
  scale_fill_identity() +
  labs(x = NULL, y = NULL) +
  theme_tufte(ticks = FALSE) +
  theme(axis.text.x = element_blank())
```

Used in anger, with grey for context and one accent for the series that carries
the argument:

```{r accent}
focus <- panel
focus$highlight <- ifelse(focus$country == "Korea", "Korea", "Other")

ggplot(focus, aes(year, trust, group = country, colour = highlight)) +
  geom_line(linewidth = 0.4) +
  geom_rangeframe(sides = "l") +
  scale_colour_manual(values = c(Korea = "#c8102e", Other = "#bfbfbf")) +
  geom_text_last(
    data = subset(focus, country == "Korea"),
    aes(label = country), size = 3, colour = "#c8102e"
  ) +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.1))) +
  labs(x = NULL, y = "Institutional trust index") +
  theme_tufte() +
  theme(legend.position = "none")
```

## Auditing the result

Every figure above can be scored. `tufte_audit()` runs fifteen checks,
structural and measured, and reports which principles the figure satisfies.

```{r audit}
final <- ggplot(experiment, aes(condition, support)) +
  geom_tufteboxplot() +
  geom_rangeframe(sides = "l") +
  labs(x = NULL, y = "Support (1-7)") +
  theme_tufte() +
  label_source("Simulated data, n = 900")

tufte_audit(final, width = 6.5, height = 4)
```

The [measuring article](measuring.html) goes through what each of those checks
is actually computing, and where the numbers can mislead you.
